---
title: "PoNet Stochastic Actor Oriented Models"
subtitle: 'This file estimates RSiena models using the PoNet data'
output:
  html_document:
    toc: true
    number_sections: true
---

\newpage

# Data prep

## Load libraries

Load libraries
```{r message = FALSE, warning = FALSE}
library(RSiena)         # siena models
library(tidyverse)      # data management
```

## Load data
```{r}
# master dataframe
load("data/df.RData")

# network data
load("data/net1.RData")
load("data/net2.RData")
load("data/net_acad.Rdata")
load("data/net_asst.Rdata")
```

# Prep RSiena objects

## Create dependent network objects

Convert friend network to dependent network variable
```{r}
# convert network to 'dependent network variable' (RSiena object)
frd <- sienaDependent(array(c(net1, net2),
                             dim = c(1584, 1584, 2)))

```

Gen dependent behavioral outcome - binary firearm discharge
```{r}
# gen outcome
gun.bin <- select(df, gun_fire_bin.1, gun_fire_bin.2)

# convert outcome into matrix
gun.bin.mat <- as.matrix(gun.bin)

# convert to rSiena object
gun.bin <- sienaDependent(gun.bin.mat, type = "behavior")
```

Gen dependent behavioral categorical - firearm use (none = 0, draw = 1, discharge = 2)
```{r}
# gen outcome
gun.cat <- select(df, gun_cat.1, gun_cat.2)

# convert outcome into matrix
gun.cat.mat <- as.matrix(gun.cat)

# convert to rSiena object
gun.cat <- sienaDependent(gun.cat.mat, type = "behavior")
```

Gen dependent behavioral outcome - 1-10+ firearm-use (draw + discharge)
```{r}
# gen outcome
gun.10 <- select(df, gun10.1, gun10.2)

# convert outcome into matrix
gun.10.mat <- as.matrix(gun.10)

# convert to RSiena object
gun.10 <- sienaDependent(gun.10.mat, type = "behavior")
```

Gen dependent behavioral outcome - quantile firearm use (draw + discharge) - 5 categories [0, 1-2, 3-4, 5-10, 11+]
```{r}
# gen outcome
gun.qt <- select(df, gun_qt.1, gun_qt.2)

# convert outcome into matrix
gun.qt.mat <- as.matrix(gun.qt)

# convert to rSiena object
gun.qt <- sienaDependent(gun.qt.mat, type = "behavior")
```

## Generate covariates

Identify constant node attributes
```{r}
# t1 value used to predict t1-t2 transition
male   <- df$male.1
whi    <- df$race_whi.1 
blk    <- df$race_blk.1
race   <- df$race
hisp   <- df$hisp.1
nyears <- df$acad_yr.1 
po     <- df$rank_po.1
pro    <- df$proactive.1  
street <- df$street.1  
threat <- df$threat.1
taser  <- df$taser.1
nops   <- df$nops          

nyears <- as.numeric(as.character(nyears))

male   <- coCovar(male)
whi    <- coCovar(whi)
blk    <- coCovar(blk)
race   <- coCovar(race)
hisp   <- coCovar(hisp)
nyears <- coCovar(nyears)
po     <- coCovar(po)
pro    <- coCovar(pro)
street <- coCovar(street)
threat <- coCovar(threat)
taser  <- coCovar(taser)
nops   <- coCovar(nops)
```

Gen constant dyadic covariates - treat as exogenous dyadic covariates
```{r}
acad <- coDyadCovar(net_acad)
asst <- coDyadCovar(net_asst)
```

Gen siena dependent file - gun.cat [firearm use (none = 0, draw = 1, discharge = 2)]
```{r}
gun.data.cat <- sienaDataCreate(frd = frd, 
                            gun     = gun.cat,
                            male    = male,
                            whi     = whi,
                            blk     = blk,
                            race    = race,
                            hisp    = hisp,
                            nyears  = nyears,
                            po      = po,
                            pro     = pro,
                            street  = street,
                            threat  = threat, 
                            taser   = taser,
                            nops    = nops,
                            acad    = acad,
                            asst    = asst)
```

Gen siena dependent file - gun.10 [1-10+ firearm-use (draw + discharge)]
```{r}
gun.data.10 <- sienaDataCreate(frd  = frd, 
                            gun     = gun.10,
                            male    = male,
                            whi     = whi,
                            blk     = blk,
                            race    = race,
                            hisp    = hisp,
                            nyears  = nyears,
                            po      = po,
                            pro     = pro,
                            street  = street,
                            threat  = threat, 
                            taser   = taser,
                            nops    = nops,
                            acad    = acad,
                            asst    = asst)
```

Gen siena dependent file - gun.qt [5 categories draw + discharge [0, 1-2, 3-4, 5-10, 11+]]
```{r}
gun.data.qt <- sienaDataCreate(frd = frd, 
                            gun     = gun.qt,
                            male    = male,
                            whi     = whi,
                            blk     = blk,
                            race    = race,
                            hisp    = hisp,
                            nyears  = nyears,
                            po      = po,
                            pro     = pro,
                            street  = street,
                            threat  = threat, 
                            taser   = taser,
                            nops    = nops,
                            acad    = acad,
                            asst    = asst)
```

Create functions to view model output
```{r}
outTable <- function(x) {
	coef <- x$theta
	coefPretty <- sprintf("%.3f", round(coef,3))
	se <- diag(x$covtheta)**.5
	sePretty <- sprintf("%.3f", round(se,3))
	pval <- 2*pnorm(-abs(coef/se))
	symp <- symnum(pval, corr = FALSE,
               cutpoints = c(0,  .001,.01,.05, .1, 1),
               symbols = c("***","**","*","."," "))
    convPretty <- sprintf("%.3f", round(abs(x$tconv),3))
    out1 <- noquote(cbind(
		Function = x$effects[[1]], 
		Effect = x$effects[[2]], 
		Coef = coefPretty, 
		StEr = sePretty, 
		Sig = symp, 
		Conv = convPretty))
	out2 <- paste("Maximum Convergence Ratio:", round(x$tconv.max,3))
	return(list(out1,out2))
}
```

\newpage

# Model specification

## Model 1a - firearm-use (none = 0, draw = 1, discharge = 2) - avAlt
gun.data.cat
```{r}
# create a model specification object for the data
myEffects1a <- getEffects(gun.data.cat)
myEffects1a

# inspect all possible effects
# write an html file to working directory
effectsDocumentation(myEffects1a) 

# test whether shared assignment and shared academy shape friendship
myEffects1a <- includeEffects(myEffects1a, X, interaction1 = "asst", name = "frd")
myEffects1a <- includeEffects(myEffects1a, X, interaction1 = "acad", name = "frd")

# test for ego effects
myEffects1a <- includeEffects(myEffects1a, egoX, interaction1 = "gun",    name = "frd")
myEffects1a <- includeEffects(myEffects1a, egoX, interaction1 = "male",   name = "frd")
myEffects1a <- includeEffects(myEffects1a, egoX, interaction1 = "nyears", name = "frd")
myEffects1a <- includeEffects(myEffects1a, egoX, interaction1 = "po",     name = "frd")
myEffects1a <- includeEffects(myEffects1a, egoX, interaction1 = "whi",    name = "frd")
myEffects1a <- includeEffects(myEffects1a, egoX, interaction1 = "blk",    name = "frd")

# test for alter effects
myEffects1a <- includeEffects(myEffects1a, altX, interaction1 = "gun",    name = "frd")
myEffects1a <- includeEffects(myEffects1a, altX, interaction1 = "male",   name = "frd")
myEffects1a <- includeEffects(myEffects1a, altX, interaction1 = "nyears", name = "frd")
myEffects1a <- includeEffects(myEffects1a, altX, interaction1 = "po",     name = "frd")
myEffects1a <- includeEffects(myEffects1a, altX, interaction1 = "whi",    name = "frd")
myEffects1a <- includeEffects(myEffects1a, altX, interaction1 = "blk",    name = "frd")

# test for homophily effects
myEffects1a <- includeEffects(myEffects1a, simX, interaction1  = "gun",    name = "frd")
myEffects1a <- includeEffects(myEffects1a, sameX, interaction1 = "male",   name = "frd")
myEffects1a <- includeEffects(myEffects1a, simX, interaction1  = "nyears", name = "frd")
myEffects1a <- includeEffects(myEffects1a, sameX, interaction1 = "po",     name = "frd")
myEffects1a <- includeEffects(myEffects1a, sameX, interaction1 = "race",   name = "frd")

# add structural effects as control variables
myEffects1a <- includeEffects(myEffects1a, density, gwespFF, gwdspFF, inPopSqrt, outActSqrt, name = "frd")

# behavioral effects - peer effects
myEffects1a <- includeEffects(myEffects1a, avAlt, name = "gun", interaction1 = "frd")

myEffects1a <- includeEffects(myEffects1a, effFrom, name = "gun", interaction1 = "male")
myEffects1a <- includeEffects(myEffects1a, effFrom, name = "gun", interaction1 = "whi")
myEffects1a <- includeEffects(myEffects1a, effFrom, name = "gun", interaction1 = "blk")
myEffects1a <- includeEffects(myEffects1a, effFrom, name = "gun", interaction1 = "hisp")
myEffects1a <- includeEffects(myEffects1a, effFrom, name = "gun", interaction1 = "nyears")
myEffects1a <- includeEffects(myEffects1a, effFrom, name = "gun", interaction1 = "po")
myEffects1a <- includeEffects(myEffects1a, effFrom, name = "gun", interaction1 = "pro")
myEffects1a <- includeEffects(myEffects1a, effFrom, name = "gun", interaction1 = "street")
myEffects1a <- includeEffects(myEffects1a, effFrom, name = "gun", interaction1 = "threat")
myEffects1a <- includeEffects(myEffects1a, effFrom, name = "gun", interaction1 = "taser")
myEffects1a <- includeEffects(myEffects1a, effFrom, name = "gun", interaction1 = "nops")

# view full model specification
myEffects1a
```

## Model 1b - firearm-use (none = 0, draw = 1, discharge = 2) - avSim
gun.data.cat
```{r}
# create a model specification object for the data
myEffects1b <- getEffects(gun.data.cat)
myEffects1b

# inspect all possible effects
# write an html file to working directory
effectsDocumentation(myEffects1b) 

# test whether shared assignment and shared academy shape friendship
myEffects1b <- includeEffects(myEffects1b, X, interaction1 = "asst", name = "frd")
myEffects1b <- includeEffects(myEffects1b, X, interaction1 = "acad", name = "frd")

# test for ego effects
myEffects1b <- includeEffects(myEffects1b, egoX, interaction1 = "gun",    name = "frd")
myEffects1b <- includeEffects(myEffects1b, egoX, interaction1 = "male",   name = "frd")
myEffects1b <- includeEffects(myEffects1b, egoX, interaction1 = "nyears", name = "frd")
myEffects1b <- includeEffects(myEffects1b, egoX, interaction1 = "po",     name = "frd")
myEffects1b <- includeEffects(myEffects1b, egoX, interaction1 = "whi",    name = "frd")
myEffects1b <- includeEffects(myEffects1b, egoX, interaction1 = "blk",    name = "frd")

# test for alter effects
myEffects1b <- includeEffects(myEffects1b, altX, interaction1 = "gun",    name = "frd")
myEffects1b <- includeEffects(myEffects1b, altX, interaction1 = "male",   name = "frd")
myEffects1b <- includeEffects(myEffects1b, altX, interaction1 = "nyears", name = "frd")
myEffects1b <- includeEffects(myEffects1b, altX, interaction1 = "po",     name = "frd")
myEffects1b <- includeEffects(myEffects1b, altX, interaction1 = "whi",    name = "frd")
myEffects1b <- includeEffects(myEffects1b, altX, interaction1 = "blk",    name = "frd")

# test for homophily effects
myEffects1b <- includeEffects(myEffects1b, simX, interaction1  = "gun",    name = "frd")
myEffects1b <- includeEffects(myEffects1b, sameX, interaction1 = "male",   name = "frd")
myEffects1b <- includeEffects(myEffects1b, simX, interaction1  = "nyears", name = "frd")
myEffects1b <- includeEffects(myEffects1b, sameX, interaction1 = "po",     name = "frd")
myEffects1b <- includeEffects(myEffects1b, sameX, interaction1 = "race",   name = "frd")

# add structural effects as control variables
myEffects1b <- includeEffects(myEffects1b, density, gwespFF, gwdspFF, inPopSqrt, outActSqrt, name = "frd")

# behavioral effects - peer effects
myEffects1b <- includeEffects(myEffects1b, avSim, name = "gun", interaction1 = "frd")

myEffects1b <- includeEffects(myEffects1b, effFrom, name = "gun", interaction1 = "male")
myEffects1b <- includeEffects(myEffects1b, effFrom, name = "gun", interaction1 = "whi")
myEffects1b <- includeEffects(myEffects1b, effFrom, name = "gun", interaction1 = "blk")
myEffects1b <- includeEffects(myEffects1b, effFrom, name = "gun", interaction1 = "hisp")
myEffects1b <- includeEffects(myEffects1b, effFrom, name = "gun", interaction1 = "nyears")
myEffects1b <- includeEffects(myEffects1b, effFrom, name = "gun", interaction1 = "po")
myEffects1b <- includeEffects(myEffects1b, effFrom, name = "gun", interaction1 = "pro")
myEffects1b <- includeEffects(myEffects1b, effFrom, name = "gun", interaction1 = "street")
myEffects1b <- includeEffects(myEffects1b, effFrom, name = "gun", interaction1 = "threat")
myEffects1b <- includeEffects(myEffects1b, effFrom, name = "gun", interaction1 = "taser")
myEffects1b <- includeEffects(myEffects1b, effFrom, name = "gun", interaction1 = "nops")

# view full model specification
myEffects1b
```

## Model 2a - firearm-use (1-10+ draw + discharge) - avAlt
gun.data.10
```{r}
# create a model specification object for the data
myEffects2a <- getEffects(gun.data.10)
myEffects2a

# inspect all possible effects
# write an html file to working directory
effectsDocumentation(myEffects2a) 

# test whether shared assignment and shared academy shape friendship
myEffects2a <- includeEffects(myEffects2a, X, interaction1 = "asst", name = "frd")
myEffects2a <- includeEffects(myEffects2a, X, interaction1 = "acad", name = "frd")

# test for ego effects
myEffects2a <- includeEffects(myEffects2a, egoX, interaction1 = "gun",    name = "frd")
myEffects2a <- includeEffects(myEffects2a, egoX, interaction1 = "male",   name = "frd")
myEffects2a <- includeEffects(myEffects2a, egoX, interaction1 = "nyears", name = "frd")
myEffects2a <- includeEffects(myEffects2a, egoX, interaction1 = "po",     name = "frd")
myEffects2a <- includeEffects(myEffects2a, egoX, interaction1 = "whi",    name = "frd")
myEffects2a <- includeEffects(myEffects2a, egoX, interaction1 = "blk",    name = "frd")

# test for alter effects
myEffects2a <- includeEffects(myEffects2a, altX, interaction1 = "gun",    name = "frd")
myEffects2a <- includeEffects(myEffects2a, altX, interaction1 = "male",   name = "frd")
myEffects2a <- includeEffects(myEffects2a, altX, interaction1 = "nyears", name = "frd")
myEffects2a <- includeEffects(myEffects2a, altX, interaction1 = "po",     name = "frd")
myEffects2a <- includeEffects(myEffects2a, altX, interaction1 = "whi",    name = "frd")
myEffects2a <- includeEffects(myEffects2a, altX, interaction1 = "blk",    name = "frd")

# test for homophily effects
myEffects2a <- includeEffects(myEffects2a, simX, interaction1  = "gun",    name = "frd")
myEffects2a <- includeEffects(myEffects2a, sameX, interaction1 = "male",   name = "frd")
myEffects2a <- includeEffects(myEffects2a, simX, interaction1  = "nyears", name = "frd")
myEffects2a <- includeEffects(myEffects2a, sameX, interaction1 = "po",     name = "frd")
myEffects2a <- includeEffects(myEffects2a, sameX, interaction1 = "race",   name = "frd")

# add structural effects as control variables
myEffects2a <- includeEffects(myEffects2a, density, gwespFF, gwdspFF, inPopSqrt, outActSqrt, name = "frd")

# behavioral effects - peer effects
myEffects2a <- includeEffects(myEffects2a, avAlt, name = "gun", interaction1 = "frd")

myEffects2a <- includeEffects(myEffects2a, effFrom, name = "gun", interaction1 = "male")
myEffects2a <- includeEffects(myEffects2a, effFrom, name = "gun", interaction1 = "whi")
myEffects2a <- includeEffects(myEffects2a, effFrom, name = "gun", interaction1 = "blk")
myEffects2a <- includeEffects(myEffects2a, effFrom, name = "gun", interaction1 = "hisp")
myEffects2a <- includeEffects(myEffects2a, effFrom, name = "gun", interaction1 = "nyears")
myEffects2a <- includeEffects(myEffects2a, effFrom, name = "gun", interaction1 = "po")
myEffects2a <- includeEffects(myEffects2a, effFrom, name = "gun", interaction1 = "pro")
myEffects2a <- includeEffects(myEffects2a, effFrom, name = "gun", interaction1 = "street")
myEffects2a <- includeEffects(myEffects2a, effFrom, name = "gun", interaction1 = "threat")
myEffects2a <- includeEffects(myEffects2a, effFrom, name = "gun", interaction1 = "taser")
myEffects2a <- includeEffects(myEffects2a, effFrom, name = "gun", interaction1 = "nops")

# view full model specification
myEffects2a
```

## Model 2b - firearm-use (1-10+ draw + discharge) - avSim
gun.data.10
```{r}
# create a model specification object for the data
myEffects2b <- getEffects(gun.data.10)
myEffects2b

# inspect all possible effects
# write an html file to working directory
effectsDocumentation(myEffects2b) 

# test whether shared assignment and shared academy shape friendship
myEffects2b <- includeEffects(myEffects2b, X, interaction1 = "asst", name = "frd")
myEffects2b <- includeEffects(myEffects2b, X, interaction1 = "acad", name = "frd")

# test for ego effects
myEffects2b <- includeEffects(myEffects2b, egoX, interaction1 = "gun",    name = "frd")
myEffects2b <- includeEffects(myEffects2b, egoX, interaction1 = "male",   name = "frd")
myEffects2b <- includeEffects(myEffects2b, egoX, interaction1 = "nyears", name = "frd")
myEffects2b <- includeEffects(myEffects2b, egoX, interaction1 = "po",     name = "frd")
myEffects2b <- includeEffects(myEffects2b, egoX, interaction1 = "whi",    name = "frd")
myEffects2b <- includeEffects(myEffects2b, egoX, interaction1 = "blk",    name = "frd")

# test for alter effects
myEffects2b <- includeEffects(myEffects2b, altX, interaction1 = "gun",    name = "frd")
myEffects2b <- includeEffects(myEffects2b, altX, interaction1 = "male",   name = "frd")
myEffects2b <- includeEffects(myEffects2b, altX, interaction1 = "nyears", name = "frd")
myEffects2b <- includeEffects(myEffects2b, altX, interaction1 = "po",     name = "frd")
myEffects2b <- includeEffects(myEffects2b, altX, interaction1 = "whi",    name = "frd")
myEffects2b <- includeEffects(myEffects2b, altX, interaction1 = "blk",    name = "frd")

# test for homophily effects
myEffects2b <- includeEffects(myEffects2b, simX, interaction1  = "gun",    name = "frd")
myEffects2b <- includeEffects(myEffects2b, sameX, interaction1 = "male",   name = "frd")
myEffects2b <- includeEffects(myEffects2b, simX, interaction1  = "nyears", name = "frd")
myEffects2b <- includeEffects(myEffects2b, sameX, interaction1 = "po",     name = "frd")
myEffects2b <- includeEffects(myEffects2b, sameX, interaction1 = "race",   name = "frd")

# add structural effects as control variables
myEffects2b <- includeEffects(myEffects2b, density, gwespFF, gwdspFF, inPopSqrt, outActSqrt, name = "frd")

# behavioral effects - peer effects
myEffects2b <- includeEffects(myEffects2b, avSim, name = "gun", interaction1 = "frd")

myEffects2b <- includeEffects(myEffects2b, effFrom, name = "gun", interaction1 = "male")
myEffects2b <- includeEffects(myEffects2b, effFrom, name = "gun", interaction1 = "whi")
myEffects2b <- includeEffects(myEffects2b, effFrom, name = "gun", interaction1 = "blk")
myEffects2b <- includeEffects(myEffects2b, effFrom, name = "gun", interaction1 = "hisp")
myEffects2b <- includeEffects(myEffects2b, effFrom, name = "gun", interaction1 = "nyears")
myEffects2b <- includeEffects(myEffects2b, effFrom, name = "gun", interaction1 = "po")
myEffects2b <- includeEffects(myEffects2b, effFrom, name = "gun", interaction1 = "pro")
myEffects2b <- includeEffects(myEffects2b, effFrom, name = "gun", interaction1 = "street")
myEffects2b <- includeEffects(myEffects2b, effFrom, name = "gun", interaction1 = "threat")
myEffects2b <- includeEffects(myEffects2b, effFrom, name = "gun", interaction1 = "taser")
myEffects2b <- includeEffects(myEffects2b, effFrom, name = "gun", interaction1 = "nops")

# view full model specification
myEffects2b
```

## Model 3a - firearm-use ([0, 1-2, 3-4, 5-10, 11+] draw + discharge) - avAlt
gun.data.qt
```{r}
# create a model specification object for the data
myEffects3a <- getEffects(gun.data.qt)
myEffects3a

# inspect all possible effects
# write an html file to working directory
effectsDocumentation(myEffects3a) 

# test whether shared assignment and shared academy shape friendship
myEffects3a <- includeEffects(myEffects3a, X, interaction1 = "asst", name = "frd")
myEffects3a <- includeEffects(myEffects3a, X, interaction1 = "acad", name = "frd")

# test for ego effects
myEffects3a <- includeEffects(myEffects3a, egoX, interaction1 = "gun",    name = "frd")
myEffects3a <- includeEffects(myEffects3a, egoX, interaction1 = "male",   name = "frd")
myEffects3a <- includeEffects(myEffects3a, egoX, interaction1 = "nyears", name = "frd")
myEffects3a <- includeEffects(myEffects3a, egoX, interaction1 = "po",     name = "frd")
myEffects3a <- includeEffects(myEffects3a, egoX, interaction1 = "whi",    name = "frd")
myEffects3a <- includeEffects(myEffects3a, egoX, interaction1 = "blk",    name = "frd")

# test for alter effects
myEffects3a <- includeEffects(myEffects3a, altX, interaction1 = "gun",    name = "frd")
myEffects3a <- includeEffects(myEffects3a, altX, interaction1 = "male",   name = "frd")
myEffects3a <- includeEffects(myEffects3a, altX, interaction1 = "nyears", name = "frd")
myEffects3a <- includeEffects(myEffects3a, altX, interaction1 = "po",     name = "frd")
myEffects3a <- includeEffects(myEffects3a, altX, interaction1 = "whi",    name = "frd")
myEffects3a <- includeEffects(myEffects3a, altX, interaction1 = "blk",    name = "frd")

# test for homophily effects
myEffects3a <- includeEffects(myEffects3a, simX, interaction1  = "gun",    name = "frd")
myEffects3a <- includeEffects(myEffects3a, sameX, interaction1 = "male",   name = "frd")
myEffects3a <- includeEffects(myEffects3a, simX, interaction1  = "nyears", name = "frd")
myEffects3a <- includeEffects(myEffects3a, sameX, interaction1 = "po",     name = "frd")
myEffects3a <- includeEffects(myEffects3a, sameX, interaction1 = "race",   name = "frd")

# add structural effects as control variables
myEffects3a <- includeEffects(myEffects3a, density, gwespFF, gwdspFF, inPopSqrt, outActSqrt, name = "frd")

# behavioral effects - peer effects
myEffects3a <- includeEffects(myEffects3a, avAlt, name = "gun", interaction1 = "frd")

myEffects3a <- includeEffects(myEffects3a, effFrom, name = "gun", interaction1 = "male")
myEffects3a <- includeEffects(myEffects3a, effFrom, name = "gun", interaction1 = "whi")
myEffects3a <- includeEffects(myEffects3a, effFrom, name = "gun", interaction1 = "blk")
myEffects3a <- includeEffects(myEffects3a, effFrom, name = "gun", interaction1 = "hisp")
myEffects3a <- includeEffects(myEffects3a, effFrom, name = "gun", interaction1 = "nyears")
myEffects3a <- includeEffects(myEffects3a, effFrom, name = "gun", interaction1 = "po")
myEffects3a <- includeEffects(myEffects3a, effFrom, name = "gun", interaction1 = "pro")
myEffects3a <- includeEffects(myEffects3a, effFrom, name = "gun", interaction1 = "street")
myEffects3a <- includeEffects(myEffects3a, effFrom, name = "gun", interaction1 = "threat")
myEffects3a <- includeEffects(myEffects3a, effFrom, name = "gun", interaction1 = "taser")
myEffects3a <- includeEffects(myEffects3a, effFrom, name = "gun", interaction1 = "nops")

# view full model specification
myEffects3a
```

## Model 3b - firearm-use ([0, 1-2, 3-4, 5-10, 11+] draw + discharge) - avSim
gun.data.qt
```{r}
# create a model specification object for the data
myEffects3b <- getEffects(gun.data.qt)
myEffects3b

# inspect all possible effects
# write an html file to working directory
effectsDocumentation(myEffects3b) 

# test whether shared assignment and shared academy shape friendship
myEffects3b <- includeEffects(myEffects3b, X, interaction1 = "asst", name = "frd")
myEffects3b <- includeEffects(myEffects3b, X, interaction1 = "acad", name = "frd")

# test for ego effects
myEffects3b <- includeEffects(myEffects3b, egoX, interaction1 = "gun",    name = "frd")
myEffects3b <- includeEffects(myEffects3b, egoX, interaction1 = "male",   name = "frd")
myEffects3b <- includeEffects(myEffects3b, egoX, interaction1 = "nyears", name = "frd")
myEffects3b <- includeEffects(myEffects3b, egoX, interaction1 = "po",     name = "frd")
myEffects3b <- includeEffects(myEffects3b, egoX, interaction1 = "whi",    name = "frd")
myEffects3b <- includeEffects(myEffects3b, egoX, interaction1 = "blk",    name = "frd")

# test for alter effects
myEffects3b <- includeEffects(myEffects3b, altX, interaction1 = "gun",    name = "frd")
myEffects3b <- includeEffects(myEffects3b, altX, interaction1 = "male",   name = "frd")
myEffects3b <- includeEffects(myEffects3b, altX, interaction1 = "nyears", name = "frd")
myEffects3b <- includeEffects(myEffects3b, altX, interaction1 = "po",     name = "frd")
myEffects3b <- includeEffects(myEffects3b, altX, interaction1 = "whi",    name = "frd")
myEffects3b <- includeEffects(myEffects3b, altX, interaction1 = "blk",    name = "frd")

# test for homophily effects
myEffects3b <- includeEffects(myEffects3b, simX, interaction1  = "gun",    name = "frd")
myEffects3b <- includeEffects(myEffects3b, sameX, interaction1 = "male",   name = "frd")
myEffects3b <- includeEffects(myEffects3b, simX, interaction1  = "nyears", name = "frd")
myEffects3b <- includeEffects(myEffects3b, sameX, interaction1 = "po",     name = "frd")
myEffects3b <- includeEffects(myEffects3b, sameX, interaction1 = "race",   name = "frd")

# add structural effects as control variables
myEffects3b <- includeEffects(myEffects3b, density, gwespFF, gwdspFF, inPopSqrt, outActSqrt, name = "frd")

# behavioral effects - peer effects
myEffects3b <- includeEffects(myEffects3b, avSim, name = "gun", interaction1 = "frd")

myEffects3b <- includeEffects(myEffects3b, effFrom, name = "gun", interaction1 = "male")
myEffects3b <- includeEffects(myEffects3b, effFrom, name = "gun", interaction1 = "whi")
myEffects3b <- includeEffects(myEffects3b, effFrom, name = "gun", interaction1 = "blk")
myEffects3b <- includeEffects(myEffects3b, effFrom, name = "gun", interaction1 = "hisp")
myEffects3b <- includeEffects(myEffects3b, effFrom, name = "gun", interaction1 = "nyears")
myEffects3b <- includeEffects(myEffects3b, effFrom, name = "gun", interaction1 = "po")
myEffects3b <- includeEffects(myEffects3b, effFrom, name = "gun", interaction1 = "pro")
myEffects3b <- includeEffects(myEffects3b, effFrom, name = "gun", interaction1 = "street")
myEffects3b <- includeEffects(myEffects3b, effFrom, name = "gun", interaction1 = "threat")
myEffects3b <- includeEffects(myEffects3b, effFrom, name = "gun", interaction1 = "taser")
myEffects3b <- includeEffects(myEffects3b, effFrom, name = "gun", interaction1 = "nops")

# view full model specification
myEffects3b
```

\newpage

# Model estimation

## Model 1a - firearm-use (none = 0, draw = 1, discharge = 2) - avAlt
```{r, eval = FALSE}
# create an object with settings to tune the estimation algorithm:
modelOptions1a <- sienaAlgorithmCreate(
	projname ='gun.data.cat', MaxDegree = c(frd = 10),  
	doubleAveraging = 0, diagonalize = .2)  

# estimate the model
myResults1a <- siena07(modelOptions1a, data = gun.data.cat,
	effects = myEffects1a, batch = FALSE, verbose = FALSE, returnDeps = TRUE
    ) 

# report the results
myResults1a         

# examine results
outTable(myResults1a)

# assign final estimates to myres
model1a <- myResults1a

# obtain results in html format
xtable(myResults1a, file="model1a.htm", type="html", digits = 3)
```

GOF Model 1a 
```{r, eval = FALSE}
# indegree - gof
( gof.id.1a <- sienaGOF(myResults1a, verbose=TRUE, varName="frd", IndegreeDistribution,
     join=T, cumulative=F) )
jpeg(file = "output/gofid1a.jpeg")
plot(gof.id.1a)
dev.off()

# outdegree - gof
( gof.od.1a <- sienaGOF(myResults1a, verbose=TRUE, varName="frd", OutdegreeDistribution,
     join=T, cumulative=F) )
jpeg(file = "output/gofod1a.jpeg")
plot(gof.od.1a)
dev.off()

# triad census - gof
( gof.tc.1a <- sienaGOF(myResults1a, verbose=TRUE, varName="frd", TriadCensus,
     join=T, levls=0:10) )
jpeg(file = "output/goftc1a.jpeg")
plot(gof.tc.1a)
dev.off()

# overall behavior distribution - gof
( gof.behaviour.1a <- sienaGOF(myResults1a, BehaviorDistribution,
	verbose=TRUE,join=TRUE,varName="gun") )
jpeg(file = "output/gofbehaviour1a.jpeg")
plot(gof.behaviour.1a)
dev.off()
```

## Model 1b - firearm-use (none = 0, draw = 1, discharge = 2) - avSim
```{r, eval = FALSE}
# Create an object with settings to tune the estimation algorithm:
modelOptions1b <- sienaAlgorithmCreate(
	projname ='gun.data.cat', MaxDegree = c(frd = 10),  
	doubleAveraging = 0, diagonalize = .2)  

# estimate the model
myResults1b <- siena07(modelOptions1b, data = gun.data.cat,
	effects = myEffects1b, batch = FALSE, verbose = FALSE, returnDeps = TRUE
    )  

# report the results
myResults1b 

# examine results
outTable(myResults1b)

# assign final estimates to myres
model1b <- myResults1b

# obtain results in html format
xtable(myResults1b, file="model1b.htm", type="html", digits = 3)
```

GOF Model 1b 
```{r, eval = FALSE}
# indegree - gof
( gof.id.1b <- sienaGOF(myResults1b, verbose=TRUE, varName="frd", IndegreeDistribution,
     join=T, cumulative=F) )
jpeg(file = "output/gofid1b.jpeg")
plot(gof.id.1b)
dev.off()

# outdegree - gof
( gof.od.1b <- sienaGOF(myResults1b, verbose=TRUE, varName="frd", OutdegreeDistribution,
     join=T, cumulative=F) )
jpeg(file = "output/gofod1b.jpeg")
plot(gof.od.1b)
dev.off()

# triad census - gof
( gof.tc.1b <- sienaGOF(myResults1b, verbose=TRUE, varName="frd", TriadCensus,
     join=T, levls=0:10) )
jpeg(file = "output/goftc1b.jpeg")
plot(gof.tc.1b)
dev.off()

# overall behavior distribution - gof
( gof.behaviour.1b <- sienaGOF(myResults1b, BehaviorDistribution,
	verbose=TRUE,join=TRUE,varName="gun") )
jpeg(file = "output/gofbehaviour1b.jpeg")
plot(gof.behaviour.1b)
dev.off()
```

## Model 2a - firearm-use (1-10+ draw + discharge) - avAlt 
gun.data.10 Does not converge
```{r, eval = FALSE}
# Create an object with settings to tune the estimation algorithm:
# modelOptions2a <- sienaAlgorithmCreate(
#	projname ='gun.data.10', MaxDegree = c(frd = 10),  
#	doubleAveraging = 0, diagonalize = .2)  


# Estimate the model
# myResults2a <- siena07(modelOptions2a, data = gun.data.10,
#	effects = myEffects2a, batch = FALSE, verbose = FALSE, returnDeps = TRUE
#    )    # if multiple processors desired, comment out this line and run the next
# , initC=T, useCluster=TRUE, nbrNodes=3) # replace "3" with your number of processors-1

# report the results
#myResults2a 

# examine fuller report of results (shows score tests)
# summary(myResults2a)

# examine results
#outTable(myResults2a)

# assign final estimates to myres
#model2a <- myResults2a

# obtain results in html format
#xtable(myResults2a, file="model2a.htm", type="html", digits = 3)
```

GOF Model 2a 
```{r, eval = FALSE}
# indegree - gof
#( gof.id.2a <- sienaGOF(myResults2a, verbose=TRUE, varName="frd", IndegreeDistribution,
#     join=T, cumulative=F) )
#jpeg(file = "output/gofid2a.jpeg")
#plot(gof.id.2a)
#dev.off()

# outdegree - gof
#( gof.od.2a <- sienaGOF(myResults2a, verbose=TRUE, varName="frd", OutdegreeDistribution,
#     join=T, cumulative=F) )
#jpeg(file = "output/gofod2a.jpeg")
#plot(gof.od.2a)
#dev.off()

# triad census - gof
#( gof.tc.2a <- sienaGOF(myResults2a, verbose=TRUE, varName="frd", TriadCensus,
#     join=T, levls=0:10) )
#jpeg(file = "output/goftc2a.jpeg")
#plot(gof.tc.2a)
#dev.off()

# goodness of fit overall behavior distribution
#( gof.behaviour.2a <- sienaGOF(myResults2a, BehaviorDistribution,
#	verbose=TRUE,join=TRUE,varName="gun") )
#jpeg(file = "output/gofbehaviour2a.jpeg")
#plot(gof.behaviour.2a)
#dev.off()
```

## Model 2b - firearm-use (1-10+ draw + discharge) - avSim
gun.data.10 Does not converge
```{r, eval = FALSE}
# Create an object with settings to tune the estimation algorithm:
#modelOptions2b <- sienaAlgorithmCreate(
#	projname ='gun.data.10', MaxDegree = c(frd = 10),  
#	doubleAveraging = 0, diagonalize = .2)  

# Estimate the model
#myResults2b <- siena07(modelOptions2b, data = gun.data.10,
#	effects = myEffects2b, batch = FALSE, verbose = FALSE, returnDeps = TRUE
#    )    # if multiple processors desired, comment out this line and run the next
# , initC=T, useCluster=TRUE, nbrNodes=3) # replace "3" with your number of processors-1

# report results
#myResults2b 

# examine fuller report of results (shows score tests)
# summary(myResults2b)

# examine results
#outTable(myResults2b)

# assign final estimates to myres
#model1 <- myResults2b

# obtain results in html format
#xtable(myResults2b, file="model2b.htm", type="html", digits = 3)
```

GOF Model 2b 
```{r, eval = FALSE}
# indegree - gof
#( gof.id.2b <- sienaGOF(myResults2b, verbose=TRUE, varName="frd", IndegreeDistribution,
#     join=T, cumulative=F) )
#jpeg(file = "output/gofid2b.jpeg")
#plot(gof.id.2b)
#dev.off()

# outdegree - gof
#( gof.od.2b <- sienaGOF(myResults2b, verbose=TRUE, varName="frd", OutdegreeDistribution,
#     join=T, cumulative=F) )
#jpeg(file = "output/gofod2b.jpeg")
#plot(gof.od.2b)
#dev.off()

# triad census - gof
#( gof.tc.2b <- sienaGOF(myResults2b, verbose=TRUE, varName="frd", TriadCensus,
#     join=T, levls=0:10) )
#jpeg(file = "output/goftc2b.jpeg")
#plot(gof.tc.2b)
#dev.off()

# overall behavior distribution - gof
#( gof.behaviour.2b <- sienaGOF(myResults2b, BehaviorDistribution,
#	verbose=TRUE,join=TRUE,varName="gun") )
#jpeg(file = "output/gofbehaviour2b.jpeg")
#plot(gof.behaviour.2b)
#dev.off()
```

## Model 3a - firearm-use ([0, 1-2, 3-4, 5-10, 11+] draw + discharge) - avAlt
```{r, eval = FALSE}
# create an object with settings to tune the estimation algorithm:
modelOptions3a <- sienaAlgorithmCreate(
	projname ='gun.data.qt', MaxDegree = c(frd = 10),  
	doubleAveraging = 0, diagonalize = .2)  

# estimate the model
myResults3a <- siena07(modelOptions3a, data = gun.data.qt,
	effects = myEffects3a, batch = FALSE, verbose = FALSE, returnDeps = TRUE
    ) 

# report results
myResults3a 

# examine results
outTable(myResults3a)

# assign final estimates to myres
model3a <- myResults3a

# obtain results in html format
xtable(myResults3a, file="model3a.htm", type="html", digits = 3)
# xtable(myAttempt2_3a, file="model3a_2.htm", type="html", digits = 3)
```

GOF Model 3a 
```{r, eval = FALSE}
# indegree - gof
( gof.id.3a <- sienaGOF(myResults3a, verbose=TRUE, varName="frd", IndegreeDistribution,
     join=T, cumulative=F) )
jpeg(file = "output/gofid3a.jpeg")
plot(gof.id.3a)
dev.off()

# outdegree - gof
( gof.od.3a <- sienaGOF(myResults3a, verbose=TRUE, varName="frd", OutdegreeDistribution,
     join=T, cumulative=F) )
jpeg(file = "output/gofod3a.jpeg")
plot(gof.od.3a)
dev.off()

# triad census - gof
( gof.tc.3a <- sienaGOF(myResults3a, verbose=TRUE, varName="frd", TriadCensus,
     join=T, levls=0:10) )
jpeg(file = "output/goftc3a.jpeg")
plot(gof.tc.3a)
dev.off()

# overall behavior distribution - gof
( gof.behaviour.3a <- sienaGOF(myResults3a, BehaviorDistribution,
	verbose=TRUE,join=TRUE,varName="gun") )
jpeg(file = "output/gofbehaviour3a.jpeg")
plot(gof.behaviour.3a)
dev.off()
```

## Model 3b - firearm-use ([0, 1-2, 3-4, 5-10, 11+] draw + discharge) - avSim
```{r, eval = FALSE}
# create an object with settings to tune the estimation algorithm:
modelOptions3b <- sienaAlgorithmCreate(
	projname ='gun.data.qt', MaxDegree = c(frd = 10),  
	doubleAveraging = 0, diagonalize = .2) 

# estimate the model
myResults3b <- siena07(modelOptions3b, data = gun.data.qt,
	effects = myEffects3b, batch = FALSE, verbose = FALSE, returnDeps = TRUE
    )  

# report the results
myResults3b    
# diagnostic checks
# If 'Overall maximum convergence ratio' is greater than .25 rerun the model
# and use previous estimates as starting values (prevAns option)
# myAttempt2_3b <- siena07(modelOptions3b, data = gun.data.qt, effects = myEffects3b, prevAns = myResults3b, returnDeps = TRUE)
# To identify troublesome effects, look for high 'Convergence t-ratios' (>.3b) 

# gen algorithm to improve model fit
 m3b_algo <- sienaAlgorithmCreate(projname = 'm3b_algo',diagonalize = 0.4, doubleAveraging = 0, n3=10000)
# (ans <- siena07(m3b_algo, data = gun.data.qt, effects = myEffects3b, prevAns = myResults3b))
myAttempt3_3b <- siena07(m3b_algo, data=gun.data.qt, effects=myEffects3b, prevAns= myResults3b, batch=FALSE, returnDeps = TRUE)

# examine fuller report of results (shows score tests)
# summary(myResults3b)
# summary(myAttempt2_3b)

# examine results
outTable(myResults3b)
# outTable(myAttempt2_3b)
outTable(myAttempt3_3b)

# assign final estimates to myres
model3b <- myResults3b

# obtain results in html format
xtable(myResults3b, file="model3b.htm", type="html", digits = 3)
xtable(myAttempt2_3b, file="model3b_2.htm", type="html", digits = 3)
xtable(myAttempt3_3b, file="model3b_3.htm", type="html", digits = 3)
```

GOF Model 3b 
```{r, eval = FALSE}
# indegree - gof
( gof.id.3b <- sienaGOF(myAttempt3_3b, verbose=TRUE, varName="frd", IndegreeDistribution,
     join=T, cumulative=F) )
jpeg(file = "output/gofid3b_2.jpeg")
plot(gof.id.3b)
dev.off()

# outdegree - gof
( gof.od.3b <- sienaGOF(myAttempt3_3b, verbose=TRUE, varName="frd", OutdegreeDistribution,
     join=T, cumulative=F) )
jpeg(file = "output/gofod3b_2.jpeg")
plot(gof.od.3b)
dev.off()

# triad census - gof
( gof.tc.3b <- sienaGOF(myAttempt3_3b, verbose=TRUE, varName="frd", TriadCensus,
     join=T, levls=0:10) )
jpeg(file = "output/goftc3b_2.jpeg")
plot(gof.tc.3b)
dev.off()

# overall behavior distribution - gof
( gof.behaviour.3b <- sienaGOF(myAttempt3_3b, BehaviorDistribution,
	verbose=TRUE,join=TRUE,varName="gun") )
jpeg(file = "output/gofbehaviour3b_2.jpeg")
plot(gof.behaviour.3b)
dev.off()
```
