---
title: "PoNet Firearm Use and Police Friendship Networks"
subtitle: 'Prepare XPD data for RSiena models, and create descriptive tables'
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 2
    toc_float: true
    df_print: paged
    theme: readable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

<!-- ------------------------------------------------------------------ -->

## Load libraries

```{r message = FALSE, warning = FALSE}
library(tidyverse)  # data management and plotting
library(lubridate)  # date helpers
library(RSiena)     # stochastic actor–oriented models
library(igraph)     # network measures
library(Hmisc)      # cut2(), categorizing data
library(psych)      # descriptives & kappa
```

---

## Read and tidy force data
```{r}
force <- read.csv("data/ponet_xpd_force19_22.csv") # force data 2019-2022
```

```{r}
# gen year variable for year force incident occurred
force$offense.date <- ymd(force$offense.date)
force$offense.year <- year(force$offense.date)

# gen firearm-variable
force <- force %>% 
  mutate(gun_force = ifelse(method.used == "firearm" |
                                method.used == "firearm/taser", 1, 0))

# gen taser-variable
force <- force %>%
  mutate(taser_force = ifelse(method.used == "taser" |
                              method.used == "asp/taser" |
                              method.used == "taser/asp" |
                              method.used == "firearm/taser" |
                              method.used == "oc/taser" |
                              method.used == "taser/oc" |
                              method.used == "physical/taser" |
                              method.used == "Physical/taser/asp" |
                              method.used == "physical/asp/taser" |
                              method.used == "physical/oc/taser", 1, 0))

# split by year and derive annual counts
force21 <- force %>% filter(offense.year == "2021")
force22 <- force %>% filter(offense.year == "2022")

# gen number force incidents annually
force21 <- force21 %>% group_by(uid) %>% mutate(nforce1 = n_distinct(incident.no, na.rm = T))
force22 <- force22 %>% group_by(uid) %>% mutate(nforce2 = n_distinct(incident.no, na.rm = T))

# gen number firearm incidents annually
force21 <- force21 %>% group_by(uid) %>% mutate(ngun_force1 = sum(gun_force, na.rm = T))
force22 <- force22 %>% group_by(uid) %>% mutate(ngun_force2 = sum(gun_force, na.rm = T))

# gen number taser incidents annually
force21 <- force21 %>% group_by(uid) %>% mutate(ntaser_force1 = sum(taser_force, na.rm = T))
force22 <- force22 %>% group_by(uid) %>% mutate(ntaser_force2 = sum(taser_force, na.rm = T))

# select relevant columns
force21 <- force21 %>% select(offense.year, uid, nforce1, ngun_force1, ntaser_force1)
force22 <- force22 %>% select(offense.year, uid, nforce2, ngun_force2, ntaser_force2)

# collapse to only include one officer per year
force21 <- force21 %>%
  group_by(uid) %>%
  slice(1) %>%
  ungroup()

force22 <- force22 %>%
  group_by(uid) %>%
  slice(1) %>%
  ungroup()
```

---

## Read and tidy OPS complaints
```{r}
# read in ops data
ops <- read.csv("data/ponet_xpd_ops.csv") # ops data

# gen year variable for year ops incident occurred (only using date from first incident within complaint)
ops$idate <- ymd(ops$idate_1)
ops$iyear <- year(ops$idate)

# remove ops incidents from 2022 >= (covariate is only measured at wave 1)
ops <- ops %>%
  filter(iyear < "2022")

# gen number ops incidents variable
ops <- ops %>%
  group_by(uid) %>%
  mutate(nops1 = n_distinct(iid))

# collapse to officer-level selecting only relevant vars for analysis
ops <- ops %>%
  select(uid, nops1) %>%
  group_by(uid) %>%
  slice(1) %>%
  ungroup()
```

Gen ops variables - up until 2022
```{r}
# read in ops data
ops22 <- read.csv("data/ponet_xpd_ops.csv") # ops data

# gen year variable for year ops incident occurred (only using date from first incident within complaint)
ops22$idate <- ymd(ops22$idate_1)
ops22$iyear <- year(ops22$idate)

# gen number ops incidents variable
ops22 <- ops22 %>%
  group_by(uid) %>%
  mutate(nops2 = n_distinct(iid))

# collapse to officer-level selecting only relevant vars for analysis
ops22 <- ops22 %>%
  select(uid, nops2) %>%
  group_by(uid) %>%
  slice(1) %>%
  ungroup()
```

---

## Import survey waves
```{r}
df21   <- read.csv("data/ponet_xpd_2021.csv")      # survey data 2021
df22   <- read.csv("data/ponet_xpd_2022.csv")      # survey data 2022
```

Clean dept asst and rank data for 2022 for officers missing this info on survey 2022 - taken from roster Nov 2022
```{r}
# if rank is missing replace w/ dept_rank
df21 <- df21 %>%
  mutate(rank = ifelse(is.na(rank), dept_rank21, rank))
df22 <- df22 %>%
  mutate(rank = ifelse(is.na(rank), dept_rank22, rank))

# if asst1 is missing replace w/ dept_asst22
df21 <- df21 %>%
  mutate(asst1 = ifelse(is.na(asst1), dept_asst21, asst1))
df22 <- df22 %>%
  mutate(asst1 = ifelse(is.na(asst1), dept_asst22, asst1))

# if street1 is missing replace w/ dept_street22
df21 <- df21 %>%
  mutate(street1 = ifelse(is.na(street1), dept_street21, street1))
df22 <- df22 %>%
  mutate(street1 = ifelse(is.na(street1), dept_street22, street1))

# if proactive1 is missing replace w/ dept_pro22
df21 <- df21 %>%
  mutate(proactive1 = ifelse(is.na(proactive1), dept_pro21, proactive1))
df22 <- df22 %>%
  mutate(proactive1 = ifelse(is.na(proactive1), dept_pro22, proactive1))
```

Merge waves and add suffixes
```{r}
df <- full_join(df21, df22, by = c("uid"))

colnames(df) <- gsub("\\.x$",'.1',colnames(df))
colnames(df) <- gsub("\\.y$",'.2',colnames(df))
```

---

## Attach force & OPS counts to survey data
```{r}
df <- left_join(df, force21, by = c("wave.1" = "offense.year", "uid" = "uid"))  # merge 2021 force data
df <- left_join(df, force22, by = c("wave.2" = "offense.year", "uid" = "uid"))  # merge 2022 force data
```

If nforce, ngun_force, or ntaser_force are missing in df, replace w/ 0 (i.e. no force reports)
```{r}
df <- df %>%
  mutate(nforce1       = ifelse(is.na(nforce1), 0, nforce1),
         nforce2       = ifelse(is.na(nforce2), 0, nforce2),
         ngun_force1   = ifelse(is.na(ngun_force1), 0, ngun_force1),
         ngun_force2   = ifelse(is.na(ngun_force2), 0, ngun_force2),
         ntaser_force1 = ifelse(is.na(ntaser_force1), 0, ntaser_force1),
         ntaser_force2 = ifelse(is.na(ntaser_force2), 0, ntaser_force2),)
```

Merge survey data and ops data
```{r}
df <- left_join(df, ops, by = c("uid"))
df <- left_join(df, ops22, by = c("uid"))
```

## Compare official records and self-reports Tables A and B

Compare firearm discharge in surveys w/ force reports - waves 1 and 2 combined. Table A
```{r}
# restructure the data into long format so can compare both waves
dfc <- df %>% select(uid, gun_fire.1, ngun_force1,
                          gun_fire.2, ngun_force2) %>%
              rename("ngun_force.1" = "ngun_force1",
                     "ngun_force.2" = "ngun_force2") %>%
              pivot_longer(
                cols = -uid,
                names_to = c(".value", "wave"),
                names_sep = "\\."
              )

# Remove rows with NA values in either variable
dfc <- dfc[!is.na(dfc$gun_fire) & !is.na(dfc$ngun_force), ]

# gen binary variables
dfc <- dfc %>%
  mutate(survey_bin = ifelse(gun_fire > 0, 1, 0),
         record_bin = ifelse(ngun_force > 0, 1, 0))

# isolate relevant vars
survey <- dfc$survey_bin
record <- dfc$record_bin 

# gen contingency table
matrix <- table(survey, record)

# print the contingency table
print(matrix)

# Calculate Accuracy
accuracy <- sum(diag(matrix)) / sum(matrix)

# Calculate Sensitivity (True Positive Rate)
sensitivity <- matrix[2, 2] / (matrix[2, 2] + matrix[1, 2])

# Calculate Specificity (True Negative Rate)
specificity <- matrix[1, 1] / (matrix[1, 1] + matrix[2, 1])

# Calculate Cohen's Kappa for agreement
library(psych)
kappa_result <- cohen.kappa(matrix)

# Output the results
cat("Confusion Matrix:\n") 
# Rows represent the self-report (0 = no, 1 = yes).
# Columns represent the official records (0 = no, 1 = yes).
print(matrix)
cat("\nAccuracy:", accuracy, "\n")
# Accuracy: Proportion of all cases that were correctly classified (both positives and negatives).
cat("Sensitivity (True Positive Rate):", sensitivity, "\n")
# Sensitivity: How well the self-report correctly identifies gun cases compared to the official record.
cat("Specificity (True Negative Rate):", specificity, "\n")
# Specificity: How well the self-report identifies non-gun cases (true negatives).
cat("Cohen's Kappa:", kappa_result$kappa, "\n")

```


Compare taser discharge in surveys w/ force reports - WAVES 1 & 2 COMBINED. Table B
```{r}
# restructure the data into long format so can compare both waves
dfc <- df %>% select(uid, taser_fire.1, ntaser_force1,
                          taser_fire.2, ntaser_force2) %>%
              rename("ntaser_force.1" = "ntaser_force1",
                     "ntaser_force.2" = "ntaser_force2") %>%
              pivot_longer(
                cols = -uid,
                names_to = c(".value", "wave"),
                names_sep = "\\."
              )

# Remove rows with NA values in either variable
dfc <- dfc[!is.na(dfc$taser_fire) & !is.na(dfc$ntaser_force), ]

# gen binary variables
dfc <- dfc %>%
  mutate(survey_bin = ifelse(taser_fire > 0, 1, 0),
         record_bin = ifelse(ntaser_force > 0, 1, 0))

# isolate relevant vars
survey <- dfc$survey_bin
record <- dfc$record_bin 

# gen contingency table
matrix <- table(survey, record)

# print the contingency table
print(matrix)

# Calculate Accuracy
accuracy <- sum(diag(matrix)) / sum(matrix)

# Calculate Sensitivity (True Positive Rate)
sensitivity <- matrix[2, 2] / (matrix[2, 2] + matrix[1, 2])

# Calculate Specificity (True Negative Rate)
specificity <- matrix[1, 1] / (matrix[1, 1] + matrix[2, 1])

# Calculate Cohen's Kappa for agreement
library(psych)
kappa_result <- cohen.kappa(matrix)

# Output the results
cat("Confusion Matrix:\n") 
print(matrix)
cat("\nAccuracy:", accuracy, "\n")
cat("Sensitivity (True Positive Rate):", sensitivity, "\n")
cat("Specificity (True Negative Rate):", specificity, "\n")
cat("Cohen's Kappa:", kappa_result$kappa, "\n")
```

---

## Generate variables for analysis

For analysis used official complaint records - Use nops1 (ops reports) unless officer missing ops reports, then use survey data
```{r}
df <- df %>%
  mutate(nops = ifelse(is.na(nops.1), nops1, nops.1))
df <- df %>%
  mutate(nops2 = ifelse(is.na(nops.2), nops2, nops.2))
```

Merge weapon data from force reports w/ survey data (i.e. if report discharging firearm in one or the other (don't sum up across))
```{r}
# take highest firearm use across force data and survey data
df <- df %>%
  mutate(gun_fire.1 = ifelse(ngun_force1 > gun_fire.1, ngun_force1, gun_fire.1),
         gun_fire.2 = ifelse(ngun_force2 > gun_fire.2, ngun_force2, gun_fire.2))

# take highest taser use across force data and survey data
df <- df %>%
  mutate(taser_fire.1 = ifelse(ntaser_force1 > taser_fire.1, ntaser_force1, taser_fire.1),
         taser_fire.2 = ifelse(ntaser_force2 > taser_fire.2, ntaser_force2, taser_fire.2))
```


Generate outcome measures
```{r}
# gen binary indicator of gun discharged past year
df <- df %>%
  mutate(gun_fire_bin.1 = ifelse(gun_fire.1 > 0, 1, 0),
         gun_fire_bin.2 = ifelse(gun_fire.2 > 0, 1, 0))

# gen continuous measure of gun use (draw + fire) wave 1 - (draw + fire + point) wave 2
df <- df %>%
  mutate(gun.1 = gun_draw.1 + gun_fire.1,
         gun.2 = gun_draw.2 + gun_fire.2)

# gen categorical variable of gun use (0 = no gun use, 1 = draw gun, 2 = fire gun)
df <- df %>%
  mutate(gun_cat.1 = ifelse(gun_fire.1 > 0, 2,
                            ifelse(gun_draw.1 > 0, 1,
                                   ifelse(gun_fire.1 == 0 & gun_draw.1 == 0, 0, NA))))
df <- df %>%
  mutate(gun_cat.2 = ifelse(gun_fire.2 > 0, 2,
                            ifelse(gun_draw.2 > 0, 1,
                                   ifelse(gun_fire.2 == 0 & gun_draw.2 == 0, 0, NA))))

# split based on equal values within groups - append gun waves 1 and 2 datasets together to determine quantiles
x <- select(df, gun.1) %>% rename("gun" = "gun.1")
y <- select(df, gun.2) %>% rename("gun" = "gun.2")
z <- rbind(x, y)

z$gun_qt <- cut2(z$gun, g = 10) # gens 5 categories (0; 1-2; 3-4; 5-10; 11-100+)

# gen dv using above values and standardizing across waves
df <- df %>%
  mutate(gun_qt.1 = ifelse(gun.1 == 0, 1,                           #[0]
                    ifelse(gun.1 >= 1  & gun.1 <= 2, 2,             #[1-2]
                    ifelse(gun.1 >= 3  & gun.1 <= 4, 3,             #[3-4]
                    ifelse(gun.1 >= 5  & gun.1 <= 10, 4,            #[5-10]
                    ifelse(gun.1 >= 11 & gun.1 <= 101, 5, NA))))))  #[11-100+]

df <- df %>%
  mutate(gun_qt.2 = ifelse(gun.2 == 0, 1,                           #[0]
                    ifelse(gun.2 >= 1  & gun.2 <= 2, 2,             #[1-2]
                    ifelse(gun.2 >= 3  & gun.2 <= 4, 3,             #[3-4]
                    ifelse(gun.2 >= 5  & gun.2 <= 10, 4,            #[5-10]
                    ifelse(gun.2 >= 11 & gun.2 <= 101, 5, NA))))))  #[11-100+]

# gen outcomes w/ 10 vars for gun-use
df <- df %>%
  mutate(gun10.1 = ifelse(gun.1 >= 10, 10, gun.1),
         gun10.2 = ifelse(gun.2 >= 10, 10, gun.2))
```

Carry forward static covariates
```{r}
df <- df %>%
  mutate(
    male.1 = case_when(
      !is.na(male.1) & !is.na(male.2) & male.1 != male.2 ~ NA_real_,
      TRUE ~ coalesce(male.1, male.2)),
    male.2 = male.1,
    hisp.1 = case_when(
      !is.na(hisp.1) & !is.na(hisp.2) & hisp.1 != hisp.2 ~ NA_real_,
      TRUE ~ coalesce(hisp.1, hisp.2)),
    hisp.2 = hisp.1,
    race.1 = case_when(!is.na(race.1) & !is.na(race.2) & race.1 != race.2 ~ NA_integer_,
      TRUE ~ coalesce(race.1, race.2))) %>%
    mutate(race.2 = race.1) %>%
    mutate(
    age.1 = coalesce(age.1, age.2 - 1),
    age.2 = coalesce(age.2, age.1 + 1),
    nyears.1 = coalesce(nyears.1, nyears.2 - 1),
    nyears.2 = coalesce(nyears.2, nyears.1 + 1),
    acad_yr.1 = coalesce(acad_yr.1, acad_yr.2),
    acad_yr.2 = coalesce(acad_yr.2, acad_yr.1),
    acad_no.1 = coalesce(acad_no.1, acad_no.2),
    acad_no.2 = coalesce(acad_no.2, acad_no.1))
```

Derived covariates
```{r}
# gen rank_po
df <- df %>%
  mutate(rank_po.1 = ifelse(rank.1 == "Police Officer" | rank.1 == "Senior Police Officer", 1, 0),
         rank_po.2 = ifelse(rank.2 == "Police Officer" | rank.1 == "Senior Police Officer", 1, 0))

# gen threat (in w1 only asked about threat, then w2 broken down by threat by weapon and threat by physical force no weapon)
df <- df %>%
  mutate(threat.1 = threat,
         threat.2 = threat_wpn + threat_phy)

# gen race
df <- df %>%
  mutate(race = ifelse(race.1 == 1, 1,                             # Black
                ifelse(race.1 == 5, 2,                             # White
                ifelse(race.1 == 2 | race.1 == 3 | race.1 == 4, 3, # Other
                       NA))))# gen race
df <- df %>%
  mutate(race = ifelse(race.1 == 1, 1,               # Black
                ifelse(race.1 == 5, 2,               # White
                ifelse(race.1 == 2 | race.1 == 3 | race.1 == 4, 3, # Other
                       NA))))

# gen race_whi
df <- df %>%
  mutate(race_whi.1 = ifelse(race.1 == 5, 1, 0),
         race_whi.2 = ifelse(race.2 == 5, 1, 0))

# gen race_blk
df <- df %>%
  mutate(race_blk.1 = ifelse(race.1 == 1, 1, 0),
         race_blk.2 = ifelse(race.2 == 1, 1, 0))

# gen race_oth
df <- df %>%
  mutate(race_oth.1 = ifelse(race.1 == 2 | race.1 == 3 | race.1 == 4, 1, 0),
         race_oth.2 = ifelse(race.2 == 2 | race.2 == 3 | race.2 == 4, 1, 0))

# gen taser use
df <- df %>%
  mutate(taser.1 = taser_draw.1 + taser_fire.1,
         taser.2 = taser_draw.2 + taser_fire.2)
```

Gen number of ties to officers outside the department
```{r}
df <- df %>%
  rowwise() %>%
  mutate(nties_outside.1 = sum(c_across(frd01.1:frd10.1) == "88888", na.rm = T),
         nties_outside.2 = sum(c_across(frd01.2:frd10.2) == "88888", na.rm = T))
```

Gen street-level variable
```{r}
# replace NA w/ 88 so ifelse statement doesn't ignore NAs
df <- df %>% 
  ungroup() %>%
  mutate(street1.1x    = ifelse(is.na(street1.1), 88, street1.1),
         street1.2x    = ifelse(is.na(street1.2), 88, street1.2),
         street2.1x    = ifelse(is.na(street2.1), 88, street2.1),
         street2.2x    = ifelse(is.na(street2.2), 88, street2.2),
         proactive1.1x = ifelse(is.na(proactive1.1), 88, proactive1.1),
         proactive1.2x = ifelse(is.na(proactive1.2), 88, proactive1.2),
         proactive2.1x = ifelse(is.na(proactive2.1), 88, proactive2.1),
         proactive2.2x = ifelse(is.na(proactive2.2), 88, proactive2.2),
         rank.1        = ifelse(is.na(rank.1), 88, rank.1),
         rank.2        = ifelse(is.na(rank.2), 88, rank.2)
         )

# gen officer street-level (and not white shirt) 
df <- df %>%
  mutate(street_asst = ifelse(street1.1x == 1 | street1.2x == 1 | street2.1x == 1 | street2.2x == 1, 1, 0),
         street_rank = ifelse(rank.1 == "Captain" | rank.1 == "Major" | rank.1 == "Deputy Chief" |
                              rank.2 == "Captain" | rank.2 == "Major" | rank.2 == "Deputy Chief", 0, 1),
         street_asst = ifelse(street1.1x == 88 & street1.2x == 88 & street2.1x == 88 & street2.2x == 88, NA, street_asst),
         street_rank = ifelse(rank.1 == 88 & rank.2 == 88, NA, street_rank),
         street = ifelse(street_asst == 1 & street_rank == 1, 1, 0))

# gen street-level based on past two years of assts
df <- df %>%
  mutate(street.1 = ifelse(street1.1x == 1 | street1.2x == 1, 1, 0),
         street.2 = ifelse(street2.1x == 1 | street2.2x == 1, 1, 0),
         street.1 = ifelse(street1.1x == 88 & street1.2x == 88, NA, street.1),
         street.2 = ifelse(street2.1x == 88 & street2.2x == 88, NA, street.2))

# gen proactive based on past two years of assts
df <- df %>%
  mutate(proactive.1 = ifelse(proactive1.1x == 1 | proactive1.2x == 1, 1, 0),
         proactive.2 = ifelse(proactive2.1x == 1 | proactive2.2x == 1, 1, 0),
         proactive.1 = ifelse(proactive1.1x == 88 & proactive1.2x == 88, NA, proactive.1),
         proactive.2 = ifelse(proactive2.1x == 88 & proactive2.2x == 88, NA, proactive.2))
```


---

## Shared academy edgelist
```{r}
# Create academy edgelist
el_acad <- df %>%
  select(acad_no.1, uid) %>%
  inner_join(., select(., acad_no.1, uid), by = "acad_no.1") %>%
  unique %>%
  arrange(uid.x, uid.y) %>%
  rename(uid = uid.x,
         alter = uid.y) %>%
  select(uid, alter, acad_no.1, everything())
```

Remove self-loops/duplicates and retain isolates - academy
```{r}
# remove all instances where uid and alter are connected based on shared NA academy no
el_acad <- el_acad %>%
  mutate(remove = ifelse(is.na(acad_no.1) & uid != alter, 1, 0)) %>%
  filter(remove != 1) %>%
  select(-c(remove))

# gen degree 
el_acad <- el_acad %>% group_by(uid) %>%
  mutate(deg = n_distinct(alter, na.rm = T) - 1)

# remove self-loops if deg > 0 (to retain isolates)
el_acad <- el_acad %>%
  mutate(remove = ifelse(uid == alter & deg > 0, 1, 0)) %>%
  filter(remove != 1) %>%
  select(-c(remove))
```

Convert edgelist to matrix for Rsiena
```{r}
# gen edge1 using cleaned edge
el_acad <- el_acad %>% select(uid, alter)

# get unique list of ALL people in the edgelist (ego and alters) - should match master network 1584 nodes
node1 <- union(unique(el_acad$uid), unique(el_acad$alter)) 

# convert to igraph object
g_acad <- igraph::graph_from_data_frame(el_acad, directed = FALSE)

# convert to matrix 
net_acad <- as.matrix(igraph::as_adjacency_matrix(g_acad))

# convert diagonal matrix to zero (remove self-loops)
diag(net_acad) <- 0

# sort rows and columns - adjacency matrix
net_acad <- net_acad[sort(rownames(net_acad)), sort(colnames(net_acad))]

# sort rows - attributes
df <- df %>%  arrange(uid)

# confirm network data and attribute data in the same order
df$uid==rownames(network::as.sociomatrix(net_acad))
```

## Shared assignment edgelist
```{r}
# reshape df for asst1-asst3 var to long format
el_asst <- df %>%
  mutate(wave = "2021") %>%
  pivot_longer(cols = asst1.1:asst3.1,
               names_to = "asst_yr",
               names_prefix = "asst",
               values_to = "asst")

# concatenate asst and asst_yr columns
el_asst$shared_asst = paste(el_asst$asst_yr, el_asst$asst, sep = "-'")

# create asst edgelist using past 3 years of assignments
el_asst <- el_asst %>%
  select(shared_asst, uid, wave) %>%
  inner_join(., select(., shared_asst, uid), by = "shared_asst") %>%
  unique %>%
  arrange(uid.x, uid.y) %>%
  rename(uid = uid.x,
         alter = uid.y) %>%
  select(uid, alter, shared_asst, everything())
```

Remove self-loops/duplicates and retain isolates - shared assignment
```{r}
# remove all instances where uid and alter are connected based on shared NA asst
el_asst <- el_asst %>%
  mutate(remove = ifelse(is.na(shared_asst) & uid != alter, 1, 0)) %>%
  filter(remove != 1) %>%
  select(-c(remove))

# gen degree 
el_asst <- el_asst %>% group_by(uid) %>%
  mutate(deg = n_distinct(alter, na.rm = T) - 1)

# remove self-loops if deg > 0 (to retain isolates)
el_asst <- el_asst %>%
  mutate(remove = ifelse(uid == alter & deg > 0, 1, 0)) %>%
  filter(remove != 1) %>%
  select(-c(remove))

# remove duplicates - instances where ego nominated same alter more than once same wave (e.g. multiple shared assignments)
el_asst <- el_asst %>%
  group_by(uid, alter) %>%
  slice(1) %>%
  ungroup()
```

Convert edgelist to matrix for Rsiena - shared assignment
```{r}
# get unique list of ALL people in the edgelist (ego and alters) - should match master network 1584 nodes
node1 <- union(unique(el_asst$uid), unique(el_asst$alter)) # 476 nodes

# convert to igraph object
g_asst <- igraph::graph_from_data_frame(el_asst, directed = FALSE)

# convert to matrix 
net_asst <- as.matrix(igraph::as_adjacency_matrix(g_asst))

# convert diagonal matrix to zero (remove self-loops)
diag(net_asst) <- 0

# sort rows and columns - adjacency matrix
net_asst <- net_asst[sort(rownames(net_asst)), sort(colnames(net_asst))]

# sort rows - attributes
df <- df %>%  arrange(uid)

# confirm network data and attribute data in the same order
df$uid==rownames(network::as.sociomatrix(net_asst))
```

---

## Friendship edgelists by wave

Need to be same dimensions for matrix transformation (keeps isolates)
```{r}
# gen wave 1 edgelist 
el1 <- df %>% 
  mutate(wave = "2021") %>%
  gather(key = "key", value = "alter", 
         frd01.1:frd10.1)  

# gen wave 2 edgelist 
el2 <- df %>% 
  mutate(wave = "2022") %>%
  gather(key = "key", value = "alter", 
         frd01.2:frd10.2)  
```

Gen edge attribute - frequency
```{r}
el1 <- el1 %>%
  mutate(frequency = 
           ifelse(key == "frd01.1", frd01_frq.1,
           ifelse(key == "frd02.1", frd02_frq.1,
           ifelse(key == "frd03.1", frd03_frq.1,
           ifelse(key == "frd04.1", frd04_frq.1,
           ifelse(key == "frd05.1", frd05_frq.1,
           ifelse(key == "frd06.1", frd06_frq.1,
           ifelse(key == "frd07.1", frd07_frq.1,
           ifelse(key == "frd08.1", frd08_frq.1,
           ifelse(key == "frd09.1", frd09_frq.1,
           ifelse(key == "frd10.1", frd10_frq.1, NA)))))))))))

el2 <- el2 %>%
  mutate(frequency = 
           ifelse(key == "frd01.2", frd01_frq.2,
           ifelse(key == "frd02.2", frd02_frq.2,
           ifelse(key == "frd03.2", frd03_frq.2,
           ifelse(key == "frd04.2", frd04_frq.2,
           ifelse(key == "frd05.2", frd05_frq.2,
           ifelse(key == "frd06.2", frd06_frq.2,
           ifelse(key == "frd07.2", frd07_frq.2,
           ifelse(key == "frd08.2", frd08_frq.2,
           ifelse(key == "frd09.2", frd09_frq.2,
           ifelse(key == "frd10.2", frd10_frq.2, NA)))))))))))


# convert frequency to numeric vector
el1 <- el1 %>%
  mutate(frequency = ifelse(frequency == "Daily", 4,
                     ifelse(frequency == "A few times a week", 3,
                     ifelse(frequency == "A few times a month", 2,
                     ifelse(frequency == "Less than once a month", 1, NA)))))

# convert frequency to numeric vector
el2 <- el2 %>%
  mutate(frequency = ifelse(frequency == "Daily", 4,
                     ifelse(frequency == "A few times a week", 3,
                     ifelse(frequency == "A few times a month", 2,
                     ifelse(frequency == "Less than once a month", 1, NA)))))
```

Gen edge attribute - years known
```{r}
el1 <- el1 %>%
  mutate(years = 
           ifelse(key == "frd01.1", frd01_yrs.1,
           ifelse(key == "frd02.1", frd02_yrs.1,
           ifelse(key == "frd03.1", frd03_yrs.1,
           ifelse(key == "frd04.1", frd04_yrs.1,
           ifelse(key == "frd05.1", frd05_yrs.1,
           ifelse(key == "frd06.1", frd06_yrs.1,
           ifelse(key == "frd07.1", frd07_yrs.1,
           ifelse(key == "frd08.1", frd08_yrs.1,
           ifelse(key == "frd09.1", frd09_yrs.1,
           ifelse(key == "frd10.1", frd10_yrs.1, NA)))))))))))

el2 <- el2 %>%
  mutate(years = 
           ifelse(key == "frd01.2", frd01_yrs.2,
           ifelse(key == "frd02.2", frd02_yrs.2,
           ifelse(key == "frd03.2", frd03_yrs.2,
           ifelse(key == "frd04.2", frd04_yrs.2,
           ifelse(key == "frd05.2", frd05_yrs.2,
           ifelse(key == "frd06.2", frd06_yrs.2,
           ifelse(key == "frd07.2", frd07_yrs.2,
           ifelse(key == "frd08.2", frd08_yrs.2,
           ifelse(key == "frd09.2", frd09_yrs.2,
           ifelse(key == "frd10.2", frd10_yrs.2, NA)))))))))))


 # replace < 1 year with '0'
el1 <- el1 %>%
  mutate(years = ifelse(years == "< 1 year", 0, years))
el2 <- el2 %>%
  mutate(years = ifelse(years == "< 1 year", 0, years))

# remove character 'years' and '<' sign from each value
el1 <- el1 %>% 
  mutate(years = str_replace_all(years, "[^[:alnum:]]", ""),
         years = str_replace_all(years, "[[A-Za-z]]", ""))
el2 <- el2 %>% 
  mutate(years = str_replace_all(years, "[^[:alnum:]]", ""),
         years = str_replace_all(years, "[[A-Za-z]]", ""))

# convert to numeric var
el1$years <- as.numeric(el1$years)
el2$years <- as.numeric(el2$years)
```

Remove self-loops/duplicates and retain isolates
```{r}
# replace 55555, 88888, and 99999 w/ NA - wave 1
el1 <- el1 %>%
  mutate(alter = ifelse(alter == "88888", NA, alter),
         alter = ifelse(alter == "99999", NA, alter),
         alter = ifelse(alter == "55555", NA, alter))

# replace 55555, 88888, and 99999 w/ NA - wave 2
el2 <- el2 %>%
  mutate(alter = ifelse(alter == "88888", NA, alter),
         alter = ifelse(alter == "99999", NA, alter),
         alter = ifelse(alter == "55555", NA, alter))

# append wave 1 and wave 2 edgelists
edge <- bind_rows(el1, el2)

# gen indicator of out-degree to retain isolates
edge <- edge %>% group_by(uid, wave) %>%
  mutate(dego = n_distinct(alter, na.rm = T))

# replace alter with uid for instances where degree == 0
edge <- edge %>%
  mutate(alter = replace(alter, dego == 0, uid))

# remove rows where alter = NA
edge <- edge %>% drop_na(alter)

# remove duplicates - instances where ego nominated same alter more than once same wave
edge <- edge %>%
  group_by(uid, alter, wave) %>%
  slice(1) %>%
  ungroup()
```

Filter edgelist to only include alters who responded to the survey in at least one wave (for some egos, none of their alters responded to the survey)
```{r}
# in el - identify if nomination responded to survey
edge$ego_survey <- edge$uid   %in% df$uid  
edge$alt_survey <- edge$alter %in% df$uid  

# calculate dego not taking into account alt_survey == FALSE (e.g. instances where none of ego's alters responded to the survey)
edge <- edge %>%
  group_by(uid, wave) %>%
  mutate(count_false = sum(alt_survey == "FALSE")) %>%
  ungroup()

# identify if none of ego's alters responded to the survey - to retain isolates
edge <- edge %>%
  mutate(alter_noresponse = ifelse(dego == count_false, 1, 0))

# if none of ego's alters responded to the survey - replace alter w/ ego's uid
edge <- edge %>%
  mutate(alter = ifelse(alter_noresponse == 1, uid, alter))

# in el - only retain nominations who responded to survey
edge <- edge %>% 
  filter(ego_survey == "TRUE") %>%
  filter(alt_survey == "TRUE" | alter_noresponse == 1)

# remove duplicates - instances where none of ego's alters responded to survey, and each alter replaced with uid
edge <- edge %>%
  group_by(uid, alter, wave) %>%
  slice(1) %>%
  ungroup()
```

Gen edgelist for 2021 and 2022 using cleaned edge
```{r}
# gen edge1 and edge2 using cleaned edge
edge1 <- edge %>% filter(wave == "2021") %>% select(uid, alter, years, frequency)
edge2 <- edge %>% filter(wave == "2022") %>% select(uid, alter, years, frequency)
```

Identify structural zeros (officers who were not employed w/ XPD in wave 1) and officers who were missing (did not take survey in wave 1)
```{r}
# identify if uid (2022) only in 2022 data (not in wave 1 survey)
link1 <- df21 %>% select(uid) %>% mutate(wave1 = 1)
edge2 <- left_join(edge2, link1, by = c("uid"))

# identify if uid (2022) only in 2022 data (not in wave 1 survey) - AND not employed by XPD in 2021
df22 <- df22 %>% mutate(acad_yr = ifelse(acad_yr == "<1981", "1980", acad_yr))
df22$acad_yr <- as.numeric(as.character(str_replace_all(df22$acad_yr, "[^[:alnum:]]", "")))
status <- df22 %>% mutate(acad_yr = 2022 - acad_yr,
                         acad_yr = ifelse(is.na(acad_yr), nyears, acad_yr)) %>%
                  select(uid, acad_yr)
edge2 <- left_join(edge2, status, by = c("uid"))
edge2 <- edge2 %>% mutate(wave1 = ifelse(is.na(wave1) & acad_yr <=2, 10, wave1))  # 2 years as this accounts for people who started training at some point in 2020 and therefore were still in training and may not have gone into in-service training until the following year

# retain relevant vars - these individuals were NA in wave 1 data
wave21_NA <- edge2 %>% filter(is.na(wave1)) %>%
  group_by(uid) %>%
  slice(1)

# retain relevant vars - these individuals were not employed in wave 1 data
wave21_struct <- edge2 %>% filter(wave1 == 10) %>%
  group_by(uid) %>%
  slice(1)
```

Identify structural zeros (officers who were no longer employed w/ XPD in wave 2) and officers who were missing (did not take survey in wave 2)
```{r}
# identify if uid (2021) only in 2021 data (not in wave 2 survey)
link2 <- df22 %>% select(uid) %>% mutate(wave2 = 1)
edge1 <- left_join(edge1, link2, by = c("uid"))

# read in employment status data - were they missing or were they not employed?
status <- read.csv("data/ponet_xpd_2022_status1.csv")
status$separation_date <- mdy(status$separation_date)
status$separation_year <- year(status$separation_date)
status <- status %>%
  mutate(status1 = ifelse(separation_year >= "2022", 10, NA)) %>%
  select(uid, status1)
status$uid <- as.numeric(as.character(status$uid))
edge1 <- left_join(edge1, status, by = c("uid"))
edge1 <- edge1 %>% mutate(wave2 = ifelse(is.na(wave2) & status1 == 10, 10, wave2))


# retain relevant vars - these individuals were NA in wave 2 data
wave22_NA <- edge1 %>% filter(is.na(wave2)) %>%
  group_by(uid) %>%
  slice(1)

# retain relevant vars - these individuals were not employed in wave 2 data
wave22_struct <- edge1 %>% filter(wave2 == 10) %>%
  group_by(uid) %>%
  slice(1)
```

Convert edgelists to matrix for Rsiena
```{r}
# get unique list of ALL people in the edgelist (ego and alters) - should match
node1 <- union(unique(edge1$uid), unique(edge1$alter)) 
node2 <- union(unique(edge2$uid), unique(edge2$alter)) 

# convert to igraph object
g1 <- igraph::graph_from_data_frame(edge1, directed = TRUE)
g2 <- igraph::graph_from_data_frame(edge2, directed = TRUE)

# convert to directed matrix 
net1 <- as.matrix(igraph::as_adjacency_matrix(g1))
net2 <- as.matrix(igraph::as_adjacency_matrix(g2)) 

# convert to directed matrix - years known
yrs1 <- as.matrix(igraph::as_adjacency_matrix(g1, attr = "years"))
yrs2 <- as.matrix(igraph::as_adjacency_matrix(g2, attr = "years"))

# convert to directed matrix - frequency
frq1 <- as.matrix(igraph::as_adjacency_matrix(g1, attr = "frequency"))
frq2 <- as.matrix(igraph::as_adjacency_matrix(g2, attr = "frequency"))

# convert diagonal matrix to zero (remove self-loops)
diag(net1) <- 0
diag(net2) <- 0
diag(yrs1) <- 0
diag(yrs2) <- 0
diag(frq1) <- 0
diag(frq2) <- 0

# sort rows and columns - adjacency matrix
net1 <- net1[sort(rownames(net1)), sort(colnames(net1))]
net2 <- net2[sort(rownames(net2)), sort(colnames(net2))]
yrs1 <- net1[sort(rownames(yrs1)), sort(colnames(yrs1))]
yrs2 <- net2[sort(rownames(yrs2)), sort(colnames(yrs2))]
frq1 <- net1[sort(rownames(frq1)), sort(colnames(frq1))]
frq2 <- net2[sort(rownames(frq2)), sort(colnames(frq2))]

# sort rows - attributes
df <- df %>%  arrange(uid)

# confirm network data and attribute data in the same order
df$uid==rownames(network::as.sociomatrix(net1))
df$uid==rownames(network::as.sociomatrix(net2))
```


Calculate jaccard index
```{r}
# Network change between subsequent observations:
(tab1to2 <- table(net1, net2,useNA='always') )
tab1to2[2,2] / (sum(tab1to2)-tab1to2[1,1])   # jaccard=.22
```

Assign structural zeros to matrix - wave 1
```{r}
table(wave21_struct$uid)
struct21 <- c("1016",	"1022",	"1098",	"1116",	"1148",	"1169",	"1175",	"1183",	"1214",	"1267",	"1342",	"1382",	"1397",	"1437",	"1447",	"1455",	"1490",	"1514",	"1566",	"1571",	"1598",	"1608",	"1613",	"1633",	"1733",	"1738",	"1756",	"1782",	"1787",	"1790",	"1792",	"1794",	"1854",	"1855",	"1857",	"1907",	"1975",	"1980",	"1990",	"1999",	"2000",	"2016",	"2032",	"2034",	"2054",	"2055",	"2071",	"2106",	"2114",	"2117",	"2127",	"2147",	"2174",	"2195",	"2211",	"2231",	"2252",	"2274",	"2275",	"2294",	"2305",	"2314",	"2316",	"2374",	"2402",	"2412",	"2490",	"2504",	"2513",	"2540",	"2547",	"2596",	"2612",	"2650",	"2652",	"2671",	"2679",	"2699",	"2710",	"2720",	"2772",	"2774",	"2783",	"2794",	"2859",	"2867",	"2881",	"2889")

# structural zeros for both sending and receiving ties
net1[struct21, ] <- 10
net1[, struct21] <- 10
```

Assign structural zeros to matrix - wave 2
```{r}
table(wave22_struct$uid)
struct22 <- c("1040",	"1049",	"1050",	"1126",	"1132",	"1204",	"1208",	"1378",	"1503",	"1582",	"1583",	"1614",	"1626",	"1681",	"1825",	"1875",	"1924",	"1955",	"1971",	"2043",	"2053",	"2063",	"2098",	"2121",	"2155",	"2184",	"2334",	"2484",	"2563",	"2601",	"2625",	"2664",	"2681",	"2692",	"2702",	"2843",	"2845",	"2880",	"2890",	"2903")

# structural zeros for both sending and receiving ties
net2[struct22, ] <- 10
net2[, struct22] <- 10
```

Assign NAs to matrix - wave 1 (n=216)
```{r}
NA21 <- c("1017",	"1032",	"1033",	"1047",	"1056",	"1071",	"1085",	"1093",	"1114",	"1133",	"1151",	"1156",	"1161",	"1172",	"1207",	"1209",	"1221",	"1223",	"1224",	"1231",	"1246",	"1250",	"1253",	"1262",	"1276",	"1277",	"1311",	"1315",	"1323",	"1329",	"1355",	"1362",	"1369",	"1375",	"1379",	"1402",	"1403",	"1404",	"1412",	"1414",	"1418",	"1431",	"1442",	"1453",	"1458",	"1464",	"1467",	"1476",	"1508",	"1511",	"1512",	"1516",	"1519",	"1520",	"1535",	"1540",	"1543",	"1562",	"1575",	"1586",	"1615",	"1617",	"1629",	"1632",	"1634",	"1635",	"1662",	"1667",	"1668",	"1669",	"1672",	"1684",	"1686",	"1690",	"1699",	"1701",	"1702",	"1719",	"1720",	"1727",	"1735",	"1740",	"1741",	"1750",	"1754",	"1759",	"1763",	"1800",	"1815",	"1829",	"1830",	"1836",	"1839",	"1849",	"1866",	"1882",	"1888",	"1909",	"1914",	"1927",	"1928",	"1931",	"1941",	"1946",	"1951",	"1953",	"1964",	"1967",	"1969",	"1989",	"1992",	"1995",	"2017",	"2027",	"2065",	"2067",	"2075",	"2076",	"2088",	"2097",	"2105",	"2128",	"2140",	"2154",	"2163",	"2165",	"2175",	"2178",	"2179",	"2194",	"2203",	"2206",	"2218",	"2220",	"2238",	"2247",	"2251",	"2260",	"2265",	"2273",	"2276",	"2284",	"2303",	"2317",	"2320",	"2325",	"2327",	"2330",	"2331",	"2346",	"2349",	"2365",	"2370",	"2399",	"2422",	"2441",	"2442",	"2461",	"2471",	"2482",	"2489",	"2495",	"2498",	"2499",	"2505",	"2528",	"2532",	"2543",	"2546",	"2550",	"2552",	"2553",	"2558",	"2561",	"2569",	"2571",	"2572",	"2595",	"2597",	"2603",	"2614",	"2637",	"2646",	"2648",	"2654",	"2659",	"2662",	"2667",	"2673",	"2676",	"2687",	"2694",	"2697",	"2713",	"2734",	"2742",	"2750",	"2759",	"2762",	"2771",	"2798",	"2822",	"2825",	"2832",	"2833",	"2836",	"2844",	"2852",	"2855",	"2857",	"2858",	"2866",	"2870",	"2897",	"2898",	"2914")

# NAs for outgoing ties only
net1[NA21, ] <- NA
```

Assign NAs to matrix - wave 2 (n=182)
```{r}
NA22 <- c("1007",	"1011",	"1018",	"1034",	"1051",	"1075",	"1080",	"1084",	"1092",	"1096",	"1103",	"1131",	"1139",	"1142",	"1180",	"1193",	"1194",	"1233",	"1251",	"1264",	"1297",	"1299",	"1309",	"1338",	"1381",	"1384",	"1387",	"1388",	"1401",	"1410",	"1417",	"1420",	"1422",	"1424",	"1432",	"1450",	"1451",	"1457",	"1463",	"1466",	"1471",	"1475",	"1485",	"1487",	"1496",	"1499",	"1517",	"1518",	"1526",	"1531",	"1538",	"1541",	"1542",	"1544",	"1546",	"1548",	"1553",	"1581",	"1587",	"1603",	"1616",	"1641",	"1646",	"1656",	"1670",	"1679",	"1685",	"1691",	"1696",	"1700",	"1705",	"1707",	"1711",	"1726",	"1729",	"1730",	"1746",	"1747",	"1766",	"1819",	"1821",	"1846",	"1853",	"1878",	"1879",	"1891",	"1902",	"1905",	"1929",	"1938",	"1948",	"1950",	"1958",	"1959",	"1974",	"1998",	"2015",	"2033",	"2036",	"2045",	"2048",	"2051",	"2113",	"2123",	"2158",	"2169",	"2187",	"2188",	"2193",	"2223",	"2224",	"2226",	"2242",	"2254",	"2255",	"2261",	"2282",	"2288",	"2295",	"2300",	"2302",	"2308",	"2309",	"2318",	"2324",	"2332",	"2344",	"2355",	"2357",	"2358",	"2379",	"2381",	"2401",	"2414",	"2416",	"2417",	"2443",	"2444",	"2447",	"2453",	"2492",	"2493",	"2514",	"2515",	"2519",	"2530",	"2541",	"2545",	"2582",	"2598",	"2631",	"2632",	"2636",	"2641",	"2642",	"2655",	"2668",	"2685",	"2701",	"2716",	"2717",	"2723",	"2725",	"2731",	"2732",	"2766",	"2768",	"2782",	"2804",	"2809",	"2812",	"2815",	"2819",	"2841",	"2853",	"2876",	"2886",	"2891",	"2896",	"2902",	"2909",	"2913")

# NAs for outgoing ties only
net2[NA22, ] <- NA
```

```{r}
table(net1, net2)
```
---

---

## Save the datasets for Rsiena models
```{r}
save(net1,     file = "data/net1.RData")
save(net2,     file = "data/net2.RData")
save(frq1,     file = "data/frq1.Rdata")
save(yrs1,     file = "data/yrs1.Rdata")
save(net_acad, file = "data/net_acad.Rdata")
save(net_asst, file = "data/net_asst.Rdata")
save(df,       file = "data/df.RData")
```

---

## Generate summary statistics for manuscript Table 1
```{r}
# load packages
library(dplyr)
library(tidyr)
library(forcats)

# build df summary
df_summary <- df %>%                     # master data frame
  select(
    uid,
    gun_qt.1,  gun_qt.2,
    nops, nops2,
    male.1, male.2,
    race_whi.1, race_whi.2, race_blk.1, race_blk.2, race_oth.1, race_oth.2,
    hisp.1,  hisp.2,
    rank_po.1, rank_po.2,
    acad_yr.1, acad_yr.2,
    proactive.1, proactive.2,
    street.1, street.2,
    threat.1, threat.2,
    taser.1,  taser.2
  ) %>%
  mutate(
    # rename counts for symmetry
    nops.1 = nops,
    nops.2 = nops2
  )

# merge scaffold into wave‑specific frames & label quartile
label_qt <- function(x) factor(x, levels = 1:5,
                               labels = c("0 times", "1–2 times",
                                          "3–4 times", "5–10 times", "11+ times"))

df_summary21 <- df21 %>%
  left_join(df_summary, by = "uid") %>%
  mutate(gun_qt_cat.1 = label_qt(gun_qt.1))

df_summary22 <- df22 %>%
  left_join(df_summary, by = "uid") %>%
  mutate(gun_qt_cat.2 = label_qt(gun_qt.2))

# gen helper functions for creating summary statistics
summarize_categorical <- function(df, var) {
  valid_df <- df %>% filter(!is.na(.data[[var]]))
  valid_n <- nrow(valid_df)
  
  valid_df %>%
    count(value = .data[[var]], name = "n") %>%
    mutate(
      variable      = var,
      category      = as.character(value),
      percent_total = round(100 * n / nrow(df), 1),
      percent_valid = round(100 * n / valid_n, 1)
    ) %>%
    select(variable, category, n, percent_total, percent_valid)
}

summarize_continuous <- function(df, var) {
  nm <- df %>% filter(!is.na(.data[[var]]))
  tibble(variable = var, category = "",
         n = nrow(nm),
         percent_total = NA, percent_valid = NA,
         mean = round(mean(nm[[var]]), 2),
         sd   = round(sd(nm[[var]]), 2),
         min  = min(nm[[var]]),
         max  = max(nm[[var]]))
}

build_summary <- function(df, wave_suffix) {
  
  # programmatic lists
  base_cat <- c("gun_qt_cat", "male", "race_whi", "race_blk", "race_oth",
                "hisp", "rank_po", "proactive", "street")
  
  base_con <- c("gun_qt", "acad_yr", "threat", "taser", "nops")
  
  cat_vars  <- paste0(base_cat, ".", wave_suffix)
  cont_vars <- paste0(base_con, ".", wave_suffix)
  
  # summaries
  cat_sum <- bind_rows(lapply(cat_vars,  summarize_categorical, df = df))
  con_sum <- bind_rows(lapply(cont_vars, summarize_continuous,  df = df))
  
  miss <- tibble(
    variable        = c(cat_vars, cont_vars),
    missing_n       = sapply(df[c(cat_vars, cont_vars)],
                             \(x) sum(is.na(x))),
    missing_percent = round(100 * sapply(df[c(cat_vars, cont_vars)],
                                         \(x) mean(is.na(x))), 1)
  )
  
  bind_rows(cat_sum, con_sum) %>%
    left_join(miss, by = "variable") %>%
    select(variable, category, n, percent_total, percent_valid,
           mean, sd, min, max, missing_n, missing_percent)
}

# gen summary stat tables for both waves
final_summary_21 <- build_summary(df_summary21, 1) %>%
  mutate(wave = "Wave 1 (2021)")

final_summary_22 <- build_summary(df_summary22, 2) %>%
  mutate(wave = "Wave 2 (2022)")

# Print or export
print(final_summary_21)
print(final_summary_22)

# optional long table
summary_combined <- bind_rows(final_summary_21, final_summary_22)

# write.csv
write.csv(final_summary_21, "output/ponet_xpd_sumstats_21.csv")
write.csv(final_summary_22, "output/ponet_xpd_sumstats_22.csv")
```

## Generate network statistics for manuscript Table 2

Calculate network statistics for each wave as presented in Table 2
```{r}
# extract relevant columns for edgelist
edge_desc1 <- select(edge1, uid, alter)
edge_desc2 <- select(edge2, uid, alter)

# create a node-level df with unique node IDs

# get unique uid
nodes_from1 <- edge_desc1 %>%
  distinct(uid) %>%
  rename(node = uid)
nodes_from2 <- edge_desc2 %>%
  distinct(uid) %>%
  rename(node = uid)

# get unique alter
nodes_to1 <- edge_desc1 %>%
  distinct(alter) %>%
  rename(node = alter)
nodes_to2 <- edge_desc2 %>%
  distinct(alter) %>%
  rename(node = alter)

# Combine them, keeping only unique rows
nodes_desc1 <- bind_rows(nodes_from1, nodes_to1) %>% distinct(node)
nodes_desc2 <- bind_rows(nodes_from2, nodes_to2) %>% distinct(node)

# merge in wave attribute
wave <- df %>% select(uid, wave.1, wave.2)
nodes_desc1 <- left_join(nodes_desc1, wave, by = c("node" = "uid"))
nodes_desc2 <- left_join(nodes_desc2, wave, by = c("node" = "uid"))
```

Gen the directed igraph object
```{r}
# convert to igraph object
g21 <- graph_from_data_frame(d = edge_desc1[, c("uid", "alter")], directed = T, vertices = nodes_desc1)
g22 <- graph_from_data_frame(d = edge_desc2[, c("uid", "alter")], directed = T, vertices = nodes_desc2)

# using wave node attribute identify if node was in wave 1 or 2 survey
v1_ids <- which(!is.na(V(g21)$wave.1) & V(g21)$wave.1 == "2021")
v2_ids <- which(!is.na(V(g22)$wave.2) & V(g22)$wave.2 == "2022")

# induce subgraph on those nodes (includes all their edges)
g21 <- induced_subgraph(g21, vids = v1_ids)
g22 <- induced_subgraph(g22, vids = v2_ids)

# simplify to remove self-loops and multiple edges
g21 <- simplify(g21, remove.multiple = TRUE, remove.loops = TRUE)
g22 <- simplify(g22, remove.multiple = TRUE, remove.loops = TRUE)
```

Calculate network measures (each wave separately)
```{r}
# Calculate network statistics - 2021
clu <- igraph::components(g21)
biggest_cluster <- which.max(clu$csize)
node_ids <- V(g21)[clu$membership == biggest_cluster]
lcc21 <- igraph::induced_subgraph(g21, node_ids)                 # gen lcc

degin21   = igraph::degree(g21, mode = "in")                     # in degree centrality for all nodes in the network
degot21   = igraph::degree(g21, mode = "out")                    # out degree centrality for all nodes in the network

V(g21)$degt21 <- igraph::degree(g21, mode = "total")             # calculate number of isolates in 2021
table(V(g21)$degt21)

network_metrics21 <- cbind(
  nodes     = vcount(g21),                                        # count the number of nodes
  edges     = ecount(g21),                                        # count the number of edges
  density   = edge_density(g21),                                  # calculate the network's density
  cc        = transitivity(g21, "global"),                        # calculate the network's clustering coefficient
  recip     = reciprocity(g21),                                   # calculate the extent to which ties in the network are reciprocated
  centr_deg = centr_degree(g21, mode = "all")$centralization,     # calculate the network's degree centralization
  lcc       = vcount(lcc21),
  meandegin = mean(degin21),
  mindegin  = min(degin21),
  maxdegin  = max(degin21),
  meandegot = mean(degot21),
  mindegot  = min(degot21),
  maxdegot  = max(degot21)
)

# convert network metrics object to a dataframe
network_metrics21 <- as.data.frame(t(network_metrics21))

# take a peak at the network statistics
options(scipen=999)
network_metrics21
```

Calculate network statistics - 2022
```{r}
clu <- igraph::components(g22)
biggest_cluster <- which.max(clu$csize)
node_ids <- V(g22)[clu$membership == biggest_cluster]
lcc22 <- igraph::induced_subgraph(g22, node_ids)                  # gen lcc

degin22   = igraph::degree(g22, mode = "in")                      # in degree centrality for all nodes in the network
degot22   = igraph::degree(g22, mode = "out")                     # out degree centrality for all nodes in the network

V(g22)$degt22 <- igraph::degree(g22, mode = "total")             # calculate number of isolates in 2022
table(V(g22)$degt22)

network_metrics22 <- cbind(
  nodes     = vcount(g22),                                        # count the number of nodes
  edges     = ecount(g22),                                        # count the number of edges
  density   = edge_density(g22),                                  # calculate the network's density
  cc        = transitivity(g22, "global"),                        # calculate the network's clustering coefficient
  recip     = reciprocity(g22),                                   # calculate the extent to which ties in the network are reciprocated
  centr_deg = centr_degree(g22, mode = "all")$centralization,     # calculate the network's degree centralization
  lcc       = vcount(lcc22),
  meandegin = mean(degin22),
  mindegin  = min(degin22),
  maxdegin  = max(degin22),
  meandegot = mean(degot22),
  mindegot  = min(degot22),
  maxdegot  = max(degot22)
)

# convert network metrics object to a dataframe
network_metrics22 <- as.data.frame(t(network_metrics22))

# take a peak at the network statistics
network_metrics22
```

Bind the network metrics
```{r}
network_metrics21 <- network_metrics21 %>% rename("2021" = "V1")
network_metrics22 <- network_metrics22 %>% rename("2022" = "V1")

network_metrics <- cbind(
  network_metrics21,
  network_metrics22
)

network_metrics

# export network statistics as csv file
write.csv(network_metrics, "output/ponet_xpd_netstats.csv")
```

---

## Generate summary statistics comparing officers in one wave versus both waves Table C 

Officers participating in both survey waves compared with one-wave participants Table C.
```{r}
# load broom package
library(broom)

# identify outdegree (replace 55555, 77777, 88888, and 99999 w/ NA)
df <- df %>%
  mutate(across(matches("^frd(0[1-9]|10)\\.(1|2)$"), 
                ~ ifelse(. %in% c("55555", "77777", "88888", "99999"), NA, .)))

df <- df %>%
    mutate(dego.1 = rowSums(!is.na(select(., frd01.1:frd10.1))),
           dego.2 = rowSums(!is.na(select(., frd01.2:frd10.2))))

# Define groups: "Both Waves" vs "One Wave"
df <- df %>% 
  mutate(group = ifelse(!is.na(wave.1) & !is.na(wave.2), "Both Waves", "One Wave"))

# transform gun_qt into dummy vars for descriptive statistics
df <- df %>%
  mutate(gun_qt_cat1.1 = ifelse(gun_qt.1 == 1, 1, 0),
         gun_qt_cat2.1 = ifelse(gun_qt.1 == 2, 1, 0),
         gun_qt_cat3.1 = ifelse(gun_qt.1 == 3, 1, 0),
         gun_qt_cat4.1 = ifelse(gun_qt.1 == 4, 1, 0),
         gun_qt_cat5.1 = ifelse(gun_qt.1 == 5, 1, 0),
         gun_qt_cat1.2 = ifelse(gun_qt.2 == 1, 1, 0),
         gun_qt_cat2.2 = ifelse(gun_qt.2 == 2, 1, 0),
         gun_qt_cat3.2 = ifelse(gun_qt.2 == 3, 1, 0),
         gun_qt_cat4.2 = ifelse(gun_qt.2 == 4, 1, 0),
         gun_qt_cat5.2 = ifelse(gun_qt.2 == 5, 1, 0),
         )

# create 3 category race variable (Blk, White and Other)
df <- df %>% 
  mutate(race3.1 = ifelse(race_blk.1 == 1, 1,
                          ifelse(race_whi.1 == 1, 2,
                                 ifelse(race_oth.1 == 1, 3, NA))))

# Use the psych::describeBy() function to get summary statistics for age and income by group
summary <- describeBy(df[, c("gun_qt_cat1.1", "gun_qt_cat2.1", "gun_qt_cat3.1", "gun_qt_cat4.1", "gun_qt_cat5.1",
                             "gun_qt_cat1.2", "gun_qt_cat2.2", "gun_qt_cat3.2", "gun_qt_cat4.2", "gun_qt_cat5.2",
                             "male.1", "race_whi.1", "race_blk.1", "race_oth.1", "hisp.1", "acad_yr.1", "rank_po.1",
                             "proactive.1", "street.1", "threat.1", "taser.1", "nops1", "dego.1", "dego.2")], group = df$group, mat = TRUE)

# Select only relevant columns (mean and standard deviation)
summary <- summary %>%
  select(group1, mean, sd)

summary <- rownames_to_column(summary, var = "Variable")

# pivot
final_s <- summary %>%
  pivot_wider(names_from = group1, values_from = c(mean, sd), names_sep = "_") %>%
  arrange(Variable)


# Perform ANOVA for the continuous variables
anova_acadyr <- aov(acad_yr.1 ~ group, data = df)
anova_threat <- aov(threat.1  ~ group, data = df)
anova_taser  <- aov(taser.1   ~ group, data = df)
anova_nops   <- aov(nops1     ~ group, data = df)
anova_dego1  <- aov(dego.1    ~ group, data = df)
anova_dego2  <- aov(dego.2    ~ group, data = df)

# Chi-squared test for the categorical variables
chi_gun1     <- chisq.test(table(df$gun_qt.1,    df$group))
chi_gun2     <- chisq.test(table(df$gun_qt.2,    df$group))
chi_male     <- chisq.test(table(df$male.1,      df$group))
chi_whi      <- chisq.test(table(df$race_whi.1,  df$group))
chi_blk      <- chisq.test(table(df$race_blk.1,  df$group))
chi_oth      <- chisq.test(table(df$race_oth.1,  df$group))
chi_race     <- chisq.test(table(df$race3.1,     df$group))
chi_hisp     <- chisq.test(table(df$hisp.1,      df$group))
chi_rankpo   <- chisq.test(table(df$rank_po.1,   df$group))
chi_pro      <- chisq.test(table(df$proactive.1, df$group))
chi_street   <- chisq.test(table(df$street.1,    df$group))

chi_gun1$p.value
chi_gun2$p.value
tidy(anova_dego1)$p.value[1]
tidy(anova_dego2)$p.value[1]
chi_male$p.value
chi_race$p.value
chi_hisp$p.value
tidy(anova_acadyr)$p.value[1]
chi_rankpo$p.value
chi_pro$p.value
chi_street$p.value
tidy(anova_threat)$p.value[1]
tidy(anova_taser)$p.value[1]
tidy(anova_nops)$p.value[1]
```



